---
title: "Predicting Daily High Temperature Using XGBoost - Summer 2026 "
subtitle: "Los Angeles International Airport (KLAX)"
author: 
  - name: "Alex Lewis"
  - name: "David Moe"
  - name: "Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction

Accurate prediction of daily high temperatures is a critical component of modern weather forecasting, influencing decision-making across sectors such as energy management, transportation, and public safety. 
Reliable temperature forecasts support efficient resource allocation, including electricity demand planning and climate-sensitive operations in aviation and infrastructure. 
In recent years, advances in data-driven modeling have complemented traditional numerical weather prediction methods, offering improved accuracy and adaptability for localized forecasting tasks [@WANG2023101499]. 
Machine learning techniques, particularly ensemble methods, have demonstrated strong performance in capturing nonlinear relationships within meteorological data [@Airlangga_2025]. 
Among these, Extreme Gradient Boosting (XGBoost) has emerged as a powerful and efficient algorithm for predictive modeling in structured datasets [@Chen_2016].

Recent research has increasingly explored the use of machine learning techniques to enhance weather prediction accuracy, particularly in short-term forecasting applications. 
Ensemble learning methods such as XGBoost have been shown to outperform traditional statistical models and complement numerical weather prediction systems by effectively capturing nonlinear relationships and correcting systematic biases [@DONG2023105579]. 
Prior studies have applied XGBoost to tasks such as precipitation forecasting and bias correction of meteorological outputs, demonstrating improved predictive performance across diverse geographic regions [@DONG2023105579]. 
Additionally, comparative analyses of machine learning algorithms highlight the robustness and computational efficiency of gradient boosting approaches when applied to structured environmental data [@Zhou_2023], [@Singh2023MachineLF], [@Airlangga_2025]. 
Despite these advances, many studies focus on regional or large-scale forecasting problems, leaving opportunities for more localized, site-specific modeling of daily temperature using high-frequency observational data. 
XGBoost has also been documented in tutorial and use-case contexts for its implementation and interpretability in applied machine learning problems [@xgboost_tutorial].

This study focuses on predicting daily high temperatures at Los Angeles International Airport (LAX) using high-frequency observational data obtained from the National Oceanic and Atmospheric Administration (NOAA) Automated Surface Observing System (ASOS) [@asos_dataset_noaa]. 
The dataset consists of five-minute interval weather measurements, providing a detailed temporal view of atmospheric conditions that can be aggregated to support daily forecasting tasks [@Khan_2025_MultiOutputTemp]. 
By leveraging this granular data, the study aims to capture short-term variability and improve the accuracy of daily maximum temperature predictions. 
An XGBoost regression model is developed to model the relationship between meteorological features and daily high temperature, with an emphasis on evaluating its predictive performance in a localized airport environment. 
This approach provides a practical framework for applying machine learning techniques to site-specific temperature forecasting using publicly available observational data.

The primary objective of this study is to develop and evaluate an XGBoost-based model for predicting daily high temperatures at LAX using aggregated ASOS observational data. 
Model performance is assessed using a comprehensive set of regression evaluation metrics, including MAE, MSE, RMSE, $R^2$, MAPE, and explained variance, to provide a thorough assessment of prediction accuracy, error magnitude, and model fit. 
In addition, the study examines feature importance to better understand the contribution of key meteorological variables to model predictions. 
By focusing on a single, well-instrumented location and leveraging high-resolution data, this research aims to demonstrate the applicability of machine learning methods for localized temperature forecasting. 
The findings contribute to the growing body of work on data-driven weather prediction and highlight the potential of XGBoost as a practical tool for operational forecasting tasks, consistent with prior statistical temperature prediction work in California and broader North America [@dias2018statistical].

## Methods

The prediction task is to forecast the daily maximum air temperature, $T_{\max}$, at weather station KLAX from surface observations. 
The relationship between the predictors and $T_{\max}$, is highly interaction‑dependent. 
For example, the cooling associated with west‑to‑east wind gusts varies by hour of the day and season, so the marginal effect of any one variable is not constant across the others. 
As a result, the conditional structure cannot be captured by additive linear terms alone. 
This analysis therefore employs gradient‑boosted regression trees as implemented in XGBoost [@Chen_2016], which are well suited to modeling nonlinear, interacting predictors in tabular meteorological data.

We construct an additive ensemble of $K$ regression trees $f_k$ drawn from the space $\mathcal{F}$,

$$
\hat{y}_i = \sum_{k=1}^{K} f_k(\mathbf{x}_i), \qquad f_k \in \mathcal{F}, \tag{1}
$$

where $x_i$ denotes the predictor vector for observation $i$ and $\hat{y}_i$ is the corresponding model prediction.
The ensemble is fitted by stagewise minimization of a regularized objective that penalizes both the number of leaves and the magnitude of leaf prediction scores in each tree:

$$
\mathcal{L} = \sum_{i} l(\hat{y}_i, y_i) + \sum_{k=1}^{K} \left[ \gamma T_k + \tfrac{1}{2}\lambda \lVert w_k \rVert^2 \right]. \tag{2}
$$
where $l(\hat{y}_i, y_i)$ is the loss function (squared error in this study), $T_k$ is the number of leaves in tree $k$, and $w_k$ is the vector of leaf scores for that tree.

This method suits our data for three reasons. 
First, it does not require the predictors to share common units or scales; variables such as wind speed and air temperature can interact naturally within the trees. 
Second, XGBoost includes explicit handling of missing values, so gaps in the observational record (e.g., sensor failures or unmeasured gusts) can be accommodated without discarding entire days or imputing values. 
Third, the fitted model can report which conditions contributed most to its predictions through feature importance and related diagnostics, allowing us to verify that the inferred drivers of high temperatures align with established meteorological cause‑and‑effect relationships.

## Analysis and Results

### Data Exploration and Visualization

#### Data Sources and Collection Process

The temperature data used in this study were obtained from the National Oceanic and Atmospheric Administration (NOAA) National Centers for Environmental Information (NCEI) Automated Surface Observing System (ASOS) network [@asos_dataset_noaa]. 
ASOS is the primary automated surface observing network in the United States and provides continuous meteorological observations at airport-based stations, including Los Angeles International Airport (KLAX), with archived one-minute and five-minute products available through NCEI. 
The dataset used here consists of 5-minute interval observations for KLAX from February 1998 through May 2026 and includes variables such as timestamp (`datetime_utc`), wind direction (`wind_dir`), wind speed (`wind_speed_kt`), wind gust (`wind_gust_kt`), visibility (`visibility_sm`), weather and sky condition fields, temperature (`temp_c`), dewpoint (`dewpoint_c`), relative humidity (`rh_pct`), and altimeter pressure (`altimeter_inhg`).

```{r, warning=FALSE, echo=T, message=FALSE}
# Setup
library(arrow)
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)
library(patchwork)
library(knitr)
library(scales)
library(tibble)
library(gtsummary)

# Data Preparation for KLAX 5-minute ASOS Data

# Inspect column names in the parquet file
parquet_path <- "KLAX_5min_complete.parquet"
df_KLAX <- read_parquet(parquet_path)

# Manually set column names
ts_col <- "datetime_utc"
temp_col <- "temp_c"

df <- df_KLAX %>%
  mutate(timestamp = as_datetime(get(ts_col), tz = "UTC"))

# Filter to last 5 years
reference_time <- ymd_hms("2026-05-01 00:00:00", tz = "UTC")

last_5_years_pre<- df %>%
  filter(timestamp >= (reference_time - years(5)))

# cat("Number of observations after filtering to last 5 years:", nrow(last_5_years_pre), "\n")

last_5_years <- last_5_years_pre %>%
  arrange(timestamp) %>%
  mutate(
    temp = get(temp_col),
    temp_change = temp - lag(temp)
  )

na_temp <- sum(is.na(last_5_years$temp_change))

# cat("Number of NA temp_change values:", sum(is.na(last_5_years$temp_change)), "\n")
# cat("Range of temp_change:", range(last_5_years$temp_change, na.rm = TRUE), "\n")

# Remove first row with NA temp_change
last_5_years <- last_5_years %>%
  filter(!is.na(temp_change))

# cat("Number of observations after removing NA temp_change:", nrow(last_5_years), "\n")

# Daily high temperature
daily_high <- last_5_years %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(date) %>%
  summarise(
    daily_high_temp = max(temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(daily_high_temp))

# cat("Number of days with daily high temperature:", nrow(daily_high), "\n")

# Diurnal data (last 2 years)
diurnal_data <- last_5_years %>%
  filter(timestamp >= (now(tzone = "UTC") - years(2))) %>%
  mutate(hour = hour(timestamp))

diurnal_avg <- diurnal_data %>%
  group_by(hour) %>%
  summarise(
    avg_temp = mean(temp, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r, warning=FALSE, echo=TRUE}
#| label: tbl-observations

n_obs_last_5_years   <- nrow(last_5_years_pre)
n_na_temp_change     <- na_temp
temp_change_range    <- range(last_5_years$temp_change, na.rm = TRUE)
n_obs_after_na_drop  <- nrow(last_5_years)
n_days_daily_high    <- nrow(daily_high)

processing_stats <- tibble(
  Statistic = c(
    "Number of observations after filtering to last 5 years",
    "Number of NA temp_change values (before removal)",
    "Minimum 5-minute temp_change (°C)",
    "Maximum 5-minute temp_change (°C)",
    "Number of observations after removing NA temp_change",
    "Number of days with daily high temperature"
  ),
  Value = c(
    n_obs_last_5_years,
    n_na_temp_change,
    temp_change_range[1],
    temp_change_range[2],
    n_obs_after_na_drop,
    n_days_daily_high
  )
)

knitr::kable(
  processing_stats,
  caption = "Observation counts and temperature-change diagnostics for the last 5 years."
)
```

For this analysis, the exploratory analysis was restricted to the most recent five years of data beginning from May 1, 2026 in order to maintain a manageable dataset size while emphasizing recent conditions most relevant to the predictive modeling task. 
After filtering, @tbl-observations shows the resulting subset contains 436,937 5-minute observations, of which 436,292 remain after removing records with missing temperature-change values, along with 1,705 days for which daily high temperature could be computed.
This subset preserves both the high temporal resolution of the original ASOS data and the seasonal coverage needed to support analysis of daily high temperature variability.

#### Descriptive Statistics

```{r, warning=FALSE, message=FALSE}
#| label: tbl-descriptive_stats

continuous_summary <- tibble(
  Variable = c(
    "Wind speed (kt)",
    "Wind gust (kt)",
    "Visibility (statute mi)",
    "Air temperature (°C)",
    "Dewpoint temperature (°C)",
    "Relative humidity (%)",
    "Altimeter (inHg)"
  ),
  N      = c(
    sum(!is.na(last_5_years$wind_speed_kt)),
    sum(!is.na(last_5_years$wind_gust_kt)),
    sum(!is.na(last_5_years$visibility_sm)),
    sum(!is.na(last_5_years$temp_c)),
    sum(!is.na(last_5_years$dewpoint_c)),
    sum(!is.na(last_5_years$rh_pct)),
    sum(!is.na(last_5_years$altimeter_inhg))
  ),
  Mean   = c(
    mean(last_5_years$wind_speed_kt,   na.rm = TRUE),
    mean(last_5_years$wind_gust_kt,    na.rm = TRUE),
    mean(last_5_years$visibility_sm,   na.rm = TRUE),
    mean(last_5_years$temp_c,          na.rm = TRUE),
    mean(last_5_years$dewpoint_c,      na.rm = TRUE),
    mean(last_5_years$rh_pct,          na.rm = TRUE),
    mean(last_5_years$altimeter_inhg,  na.rm = TRUE)
  ),
  SD     = c(
    sd(last_5_years$wind_speed_kt,     na.rm = TRUE),
    sd(last_5_years$wind_gust_kt,      na.rm = TRUE),
    sd(last_5_years$visibility_sm,     na.rm = TRUE),
    sd(last_5_years$temp_c,            na.rm = TRUE),
    sd(last_5_years$dewpoint_c,        na.rm = TRUE),
    sd(last_5_years$rh_pct,            na.rm = TRUE),
    sd(last_5_years$altimeter_inhg,    na.rm = TRUE)
  ),
 Q1     = c(
    quantile(last_5_years$wind_speed_kt,   0.25, na.rm = TRUE),
    quantile(last_5_years$wind_gust_kt,    0.25, na.rm = TRUE),
    quantile(last_5_years$visibility_sm,   0.25, na.rm = TRUE),
    quantile(last_5_years$temp_c,          0.25, na.rm = TRUE),
    quantile(last_5_years$dewpoint_c,      0.25, na.rm = TRUE),
    quantile(last_5_years$rh_pct,          0.25, na.rm = TRUE),
    quantile(last_5_years$altimeter_inhg,  0.25, na.rm = TRUE)
  ),
  Median = c(
    median(last_5_years$wind_speed_kt,    na.rm = TRUE),
    median(last_5_years$wind_gust_kt,     na.rm = TRUE),
    median(last_5_years$visibility_sm,    na.rm = TRUE),
    median(last_5_years$temp_c,           na.rm = TRUE),
    median(last_5_years$dewpoint_c,       na.rm = TRUE),
    median(last_5_years$rh_pct,           na.rm = TRUE),
    median(last_5_years$altimeter_inhg,   na.rm = TRUE)
  ),
  Q3     = c(
    quantile(last_5_years$wind_speed_kt,   0.75, na.rm = TRUE),
    quantile(last_5_years$wind_gust_kt,    0.75, na.rm = TRUE),
    quantile(last_5_years$visibility_sm,   0.75, na.rm = TRUE),
    quantile(last_5_years$temp_c,          0.75, na.rm = TRUE),
    quantile(last_5_years$dewpoint_c,      0.75, na.rm = TRUE),
    quantile(last_5_years$rh_pct,          0.75, na.rm = TRUE),
    quantile(last_5_years$altimeter_inhg,  0.75, na.rm = TRUE)
  ),
  Min    = c(
    min(last_5_years$wind_speed_kt,    na.rm = TRUE),
    min(last_5_years$wind_gust_kt,     na.rm = TRUE),
    min(last_5_years$visibility_sm,    na.rm = TRUE),
    min(last_5_years$temp_c,          na.rm = TRUE),
    min(last_5_years$dewpoint_c,      na.rm = TRUE),
    min(last_5_years$rh_pct,          na.rm = TRUE),
    min(last_5_years$altimeter_inhg,  na.rm = TRUE)
  ),
  Max    = c(
    max(last_5_years$wind_speed_kt,    na.rm = TRUE),
    max(last_5_years$wind_gust_kt,     na.rm = TRUE),
    max(last_5_years$visibility_sm,    na.rm = TRUE),
    max(last_5_years$temp_c,           na.rm = TRUE),
    max(last_5_years$dewpoint_c,       na.rm = TRUE),
    max(last_5_years$rh_pct,           na.rm = TRUE),
    max(last_5_years$altimeter_inhg,   na.rm = TRUE)
  )
)

knitr::kable(
  continuous_summary,
  caption = "Descriptive statistics for KLAX 5-minute meteorological variables over the last five years.",
  digits = 2
)
```

The descriptive statistics of the selected continuous variables in @tbl-descriptive_stats indicate that most predictors are well-behaved but not strictly Gaussian, with moderate skewness, truncation, and differing scales across variables, which supports the use of a flexible, nonparametric model such as XGBoost. 
Wind speed and gusts show right-tailed behavior with occasional stronger events, while relative humidity exhibits a broader spread and occasional very dry conditions. 
Visibility is heavily concentrated at its upper bound (10 statute miles), creating a pronounced ceiling effect, and altimeter readings are tightly clustered with minimal variance. 
In combination, these features form a predictor set with mixed distributional shapes, bounded ranges, and differing levels of measurement frequency, such as the smaller number of observations for gusts, making tree-based gradient boosting particularly suitable because it does not rely on linearity or normality assumptions and handles non-linear interactions and threshold effects naturally.

From a preprocessing standpoint, the ranges and standard deviations suggest that standard feature scaling or normalization can be used to stabilize training without requiring aggressive transformations of each variable, further aligning with standard XGBoost workflows. Variables such as wind speed, dewpoint temperature, and relative humidity carry substantial variability and are likely to provide strong predictive signal. XGBoost can also accommodate the reduced sample size for wind gusts, by learning optimal directions for missing values in the tree structure rather than requiring imputation strategies. Taken together, these descriptive statistics justify the choice of XGBoost as an appropriate modeling framework for this dataset, as it can exploit non-linear relationships and interaction effects while remaining robust to skewed, bounded, and partially observed variables.

#### Initial Findings and Insights

Exploratory analysis of the filtered KLAX dataset reveals a clear and persistent annual cycle in daily high temperature. 
The time series of daily high temperatures over the last five years in @fig-daily-high-ts shows regular seasonal oscillation, with warmer peaks in late summer and cooler conditions in winter, while still exhibiting day-to-day fluctuations and occasional extreme warm events.

```{r, warning=FALSE, echo=TRUE}
#| label: fig-daily-high-ts
#| fig-cap: "Daily high temperature at KLAX over the last five years."

p_daily_high <- ggplot(daily_high, aes(x = date, y = daily_high_temp)) +
  geom_line(color = "darkorange", alpha = 0.8) +
  labs(
    title = "Daily High Temperature at KLAX (Last 5 Years)",
    x = "Date",
    y = "Daily High Temperature (°C)"
  ) +
  theme_minimal() +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y")

p_daily_high

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_daily_high_temp_ts.png"),
  plot = p_daily_high,
  width = 7, height = 5, dpi = 300
)
```

This seasonal structure is reinforced by the monthly boxplots in @fig-monthly-boxplot, which show median daily highs increasing gradually from winter through late summer and declining again through autumn, with August and September containing the highest medians and some of the warmest outlying days in the sample.

```{r, warning=FALSE, echo=TRUE}
#| label: fig-monthly-boxplot
#| fig-cap: "Monthly distribution of daily high temperature at KLAX over the last five years. Median temperatures rise from winter to late summer and decline through autumn."

daily_high_months <- daily_high %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE))

p_boxplot <- ggplot(daily_high_months, aes(x = month, y = daily_high_temp, fill = month)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  labs(
    title = "Monthly Distribution of Daily High Temperature at KLAX",
    x = "Month",
    y = "Daily High Temperature (°C)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")

p_boxplot

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_boxplot.png"),
  plot = p_boxplot,
  width = 7, height = 5, dpi = 300
)
```

@fig-daily-high-ranges below shows the same seasonal structure at finer resolution, with mean daily highs climbing steadily from January into late summer, peaking in early September, and then declining through autumn into December. The shaded band and dashed lines highlight a wider spread between the daily min and max during the cooler winter season, indicating that while the average highs are higher in late summer, the cooler individual winter days actually exhibit greater variability and more frequent extremes than in the warmer months.

```{r, warning=FALSE, message=FALSE}
#| label: fig-daily-high-ranges
#| fig-cap: "Daily high temperature at KLAX over the annual cycle, showing daily min, max, mean, and interquartile range values over the last five years. Mean highs increase from winter into late summer, peak in early September, and then decline through autumn."

# Add year and day-of-year
daily_high_doy <- daily_high %>%
  mutate(
    year = year(date),
    doy  = yday(date)
  ) %>%
  filter(doy <= 365)

# Compute daily statistics across the last 5 years
doy_stats <- daily_high_doy %>%
  group_by(doy) %>%
  summarise(
    mean_high = mean(daily_high_temp, na.rm = TRUE),
    q1_high   = quantile(daily_high_temp, 0.25, na.rm = TRUE),
    q3_high   = quantile(daily_high_temp, 0.75, na.rm = TRUE),
    min_high  = min(daily_high_temp, na.rm = TRUE),
    max_high  = max(daily_high_temp, na.rm = TRUE),
    .groups   = "drop"
  )

# Approximate month positions for labeling
month_starts <- tibble(
  doy = c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
  month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
)

p_daily_high_ranges <- ggplot(doy_stats, aes(x = doy)) +
  # Interquartile range ribbon
  geom_ribbon(
    aes(ymin = q1_high, ymax = q3_high),
    fill = "skyblue3",
    alpha = 0.35
  ) +
  # Min/max envelope lines 
  geom_line(
    aes(y = min_high, color = "Min"),
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = max_high, color = "Max"),
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  # Mean daily high line 
  geom_line(
    aes(y = mean_high, color = "Mean"),
    linewidth = 0.9
  ) +
  scale_color_manual(
    name = "Statistic: ",
    values = c(
      "Min"  = "gray50",
      "Max"  = "gray50",
      "Mean" = "firebrick3"
    )
  ) +
  scale_x_continuous(
    breaks = month_starts$doy,
    labels = month_starts$month,
    expand = c(0.01, 0.01)
  ) +
  labs(
    title = "Daily High Temperature at KLAX by Day of Year",
    x = "Month",
    y = "Daily high temperature (°C)"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor  = element_blank(),
    legend.position   = "bottom"
  )

p_daily_high_ranges

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_daily_high_ranges.png"),
  plot = p_daily_high_ranges,
  width = 7, height = 5, dpi = 300
)
```

The overall distribution of daily high temperatures in @fig-daily-high-hist appears to be normally distributed and concentrated within a moderate range, with most days falling between approximately 16 °C and 24 °C and relatively few very cool or very hot days. 

```{r, warning=FALSE, echo=TRUE}
#| label: fig-daily-high-hist
#| fig-cap: "Histogram of daily high temperatures at KLAX over the last five years. Most daily highs fall within a moderate range, while very hot and very cool days occur less frequently."

p_hist_daily <- ggplot(daily_high, aes(x = daily_high_temp)) +
  geom_histogram(binwidth = 1, fill = "forestgreen", color = "white", alpha = 0.8) +
  labs(
    title = "Histogram of Daily High Temperature at KLAX (Last 5 Years)",
    x = "Daily High Temperature (°C)",
    y = "Frequency"
  ) +
  theme_minimal()

p_hist_daily

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_hist_daily_high_temps.png"),
  plot = p_hist_daily,
  width = 7, height = 5, dpi = 300
)
```

Together, @fig-daily-high-ts, @fig-monthly-boxplot, and @fig-daily-high-hist show that the dataset contains strong seasonal signal without being dominated by erratic or implausible daily values, making it suitable for subsequent machine learning analysis.

At shorter timescales, the five-minute observations exhibit strong persistence. 
The histogram of all 5-minute temperature changes in @fig-temp-change-all shows an overwhelming concentration at 0 °C, and the non-zero histogram in @fig-temp-change-nonzero confirms that even when change occurs, most changes are small in magnitude and cluster tightly around zero. 

```{r, warning=FALSE, echo=TRUE}
#| label: fig-temp-change-all
#| fig-cap: "Histogram of all 5-minute temperature changes at KLAX over the last five years. The distribution is strongly concentrated at 0 °C, indicating that most short-term intervals show no measurable temperature change."

p_hist_change <- ggplot(last_5_years, aes(x = temp_change)) +
  geom_histogram(
    binwidth = 0.1,
    fill = "steelblue",
    color = "white",
    alpha = 0.8
  ) +
  scale_x_continuous(breaks = seq(-2, 2, by = 0.5)) +
  labs(
    title = "Histogram of 5-Minute Temperature Changes at KLAX",
    subtitle = "Most 5-minute changes are exactly 0°C; extreme changes are rare outliers",
    x = "5-Minute Temperature Change (°C)",
    y = "Frequency"
  ) +
  theme_minimal() +
  xlim(-3, 3)

p_hist_change

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_5min_temp_change_hist.png"),
  plot = p_hist_change,
  width = 7, height = 5, dpi = 300
)
```

```{r, warning=FALSE, echo=TRUE}
#| label: fig-temp-change-nonzero
#| fig-cap: "Histogram of non-zero 5-minute temperature changes at KLAX over the last five years. Most non-zero changes remain small in magnitude, with only a small number of larger positive or negative jumps."

# Proportion of zero changes and histogram of non-zero changes
zero_changes <- last_5_years %>%
  filter(temp_change == 0)

nonzero_changes <- last_5_years %>%
  filter(temp_change != 0)

prop_zero <- nrow(zero_changes) / nrow(last_5_years)

# cat("Proportion of 5-minute intervals with exactly 0°C change:", prop_zero, "\n")
# cat("Percentage:", 100 * prop_zero, "%\n")

p_non_zero_hist <- ggplot(nonzero_changes, aes(x = temp_change)) +
  geom_histogram(
    binwidth = 0.2,
    fill = "darkorange",
    color = "white",
    alpha = 0.8
  ) +
  labs(
    title = "Histogram of Non-Zero 5-Minute Temperature Changes at KLAX",
    subtitle = sprintf("Proportion of zero changes: %.1f%%", 100 * prop_zero),
    x = "5-Minute Temperature Change (°C)",
    y = "Frequency"
  ) +
  theme_minimal() +
  xlim(-5, 5)

p_non_zero_hist

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_5min_nonzero_change_hist.png"),
  plot = p_non_zero_hist,
  width = 7, height = 5, dpi = 300
)
```

This pattern is consistent with the summary statistics in @tbl-temp-summary: approximately 70.76% of five-minute intervals show exactly 0 °C change, the mean absolute five-minute change is 0.17 °C, the median absolute change is 0 °C, and the 75th percentile is only 0.5 °C, indicating that short-term temperature changes at KLAX are typically gradual rather than abrupt.

```{r, warning=FALSE, echo=TRUE}
#| label: fig-3h-temp-change
#| fig-cap: "Distribution of 3-hour temperature changes from 5-minute observations, 2000-2026"

d3 <- df_KLAX

d3 <- d3 %>%
  mutate(t3 = lag(datetime_utc, 36),
         temp_lag = lag(temp_c, 36),
         delta = temp_c - temp_lag) %>%
  filter(as.numeric(datetime_utc - t3, units = "mins") == 180, !is.na(delta))

h3_temp_change <- ggplot(d3, aes(delta)) +
  geom_histogram(binwidth = 1, boundary = 0, fill = "steelblue", colour = "white") +
  scale_x_continuous(breaks = seq(-10, 10, by = 2)) +
  coord_cartesian(xlim = c(-10, 10)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Histogram of 3-hour Temperature Changes at LAX",
    subtitle = "Distribution of 3-hour temperature changes from 5-minute observations, 2000-2026",
    x = "Temperature change over 3 hours (°C)",
    y = "Number of 3-hour periods",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    plot.caption = element_text(size = 8, colour = "grey50")
  )

h3_temp_change

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_3h_temp_change.png"),
  plot = h3_temp_change,
  width = 7, height = 5, dpi = 300
)
```

```{r, warning=FALSE, echo=TRUE}
#| label: fig-3h-wind-change
#| fig-cap: "3-hour temperature change following gust observations, 2000-2026"

g <- df_KLAX

g <- g %>%
  mutate(temp_future = lead(temp_c, 36),
         t_future = lead(datetime_utc, 36),
         dT = temp_future - temp_c,
         wd = as.numeric(wind_dir)) %>%
  filter(!is.na(wind_gust_kt), wind_gust_kt < 100,
         !is.na(wd), !is.na(dT),
         as.numeric(t_future - datetime_utc, units = "mins") == 180,
         abs(dT) <= 10) %>%
  mutate(group = case_when(
    wd >= 210 & wd <= 330 ~ "Westerly gusts (west to east)",
    wd >= 30  & wd <= 150 ~ "Easterly gusts (east to west)",
    TRUE ~ NA_character_)) %>%
  filter(!is.na(group))

h3_wind_change <- ggplot(g, aes(dT, fill = group)) +
  geom_histogram(binwidth = 1, boundary = 0, position = "identity",
                 alpha = 0.6, colour = "white") +
  scale_x_continuous(breaks = seq(-10, 10, by = 2)) +
  coord_cartesian(xlim = c(-10, 10)) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Westerly gusts (west to east)" = "lightblue",
                               "Easterly gusts (east to west)" = "red")) +
  labs(title = "Temperature Change After Gusts at LAX: Onshore vs Offshore",
       subtitle = "3-hour temperature change following gust observations, 2000-2026",
       x = "Temperature change over 3 hours (°C)",
       y = "Number of gust observations",
       fill = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 10, colour = "grey40"),
        plot.caption = element_text(size = 8, colour = "grey50"),
        legend.position = "top")

h3_wind_change

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_3h_wind_change.png"),
  plot = h3_wind_change,
  width = 7, height = 5, dpi = 300
)
```

```{r, warning=FALSE, echo=TRUE}
#| label: tbl-temp-summary

# Summary statistics and abnormal change flags for 5-minute temperature changes

# Absolute temperature changes
last_5_years <- last_5_years %>%
  mutate(abs_temp_change = abs(temp_change))

# Quantiles of absolute changes
abs_q <- quantile(last_5_years$abs_temp_change,
                  probs = c(0.5, 0.75, 0.9, 0.95, 0.99, 1.0))

# Define abnormal threshold (99th percentile)
abnormal_threshold <- abs_q["99%"]

# Flag abnormal changes
last_5_years <- last_5_years %>%
  mutate(abnormal_change = abs_temp_change > abnormal_threshold)

n_abnormal <- sum(last_5_years$abnormal_change)
n_total <- nrow(last_5_years)
pct_abnormal <- 100 * n_abnormal / n_total

# Summary statistics table
summary_stats <- tibble(
  Statistic = c(
    "Total observations",
    "Mean absolute 5-min change (°C)",
    "Median absolute 5-min change (°C)",
    "75th percentile (°C)",
    "90th percentile (°C)",
    "95th percentile (°C)",
    "99th percentile (°C) [abnormal threshold]",
    "Maximum absolute 5-min change (°C)",
    "Number of abnormal changes",
    "Percentage of abnormal changes (%)"
  ),
  Value_raw = c(
    n_total,
    mean(last_5_years$abs_temp_change, na.rm = TRUE),
    median(last_5_years$abs_temp_change, na.rm = TRUE),
    abs_q["75%"],
    abs_q["90%"],
    abs_q["95%"],
    abs_q["99%"],
    max(last_5_years$abs_temp_change, na.rm = TRUE),
    n_abnormal,
    pct_abnormal
  )
) %>%
  mutate(
    Value = case_when(
      Statistic %in% c("Total observations", "Number of abnormal changes") ~
        comma(Value_raw, accuracy = 1),
      Statistic == "Percentage of abnormal changes (%)" ~
        number(Value_raw, accuracy = 0.01),
      TRUE ~ number(Value_raw, accuracy = 0.01)
    )
  )

knitr::kable(
  summary_stats %>% select(Statistic, Value),
  caption = "Summary statistics for 5-minute temperature changes."
)
```

The diurnal cycle plot in @fig-diurnal-cycle provides an additional view of the short-term structure by averaging temperature across each hour of the day during the last two years. That figure shows coolest temperatures during the early morning hours, a steady increase through the morning, and a peak in the early afternoon before gradual evening cooling, which is consistent with expected solar-heating dynamics. This smooth and interpretable daily pattern suggests that time-based predictors such as hour, day, or season are likely to be useful in downstream modeling.

```{r, warning=FALSE, echo=TRUE}
#| label: fig-diurnal-cycle
#| fig-cap: "Average diurnal temperature cycle at KLAX based on the last two years of 5-minute observations. Temperatures are coolest in the early morning, rise through the morning hours, peak in the early afternoon, and then gradually decline into the evening."

p_diurnal <- ggplot(diurnal_avg, aes(x = hour, y = avg_temp)) +
  geom_line(color = "purple", size = 1.2) +
  geom_point(color = "purple", size = 2) +
  labs(
    title = "Average Diurnal Temperature Curve at KLAX (Last 2 Years)",
    x = "Hour of Day",
    y = "Average Temperature (°C)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 0:23)

p_diurnal

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

ggsave(
  filename = file.path(output_dir, "fig_diurnal_cycle.png"),
  plot = p_diurnal,
  width = 7, height = 5, dpi = 300
)
```

#### Unexpected Patterns and Anomalies

Although most short-term temperature changes are small, the exploratory analysis also reveals a small number of unusually large jumps. 
Based on the absolute change distribution shown in @fig-temp-change-all and @fig-temp-change-nonzero, the 90th and 95th percentiles are approximately 0.6 °C, while the 99th percentile is 1.1 °C and can be used as a practical threshold for identifying abnormal five-minute changes. Only 4,332 observations, or about 0.99% of the five-minute records, exceed this threshold, indicating that abrupt short-term temperature jumps are rare in the dataset overall.

The largest observed absolute five-minute change reaches 22.8 °C, which is physically extreme for such a short interval and likely reflects either a reporting artifact, a timing inconsistency, or another unusual observational circumstance rather than typical atmospheric behavior. 
Even so, the rarity of these values suggests that they do not dominate the series, but they do warrant attention during preprocessing because they may affect derived features or distort model fitting if left untreated. 
Furthermore, the presence of long stretches of unchanged temperature and a small number of large jumps indicates that a more practical approach of analysis could be to use a larger timescale.

A more detailed example of short-term multi-variable structure is shown in @fig-multivariable-window, which presents a 36-hour sequence of 5-minute KLAX observations from July 22 to July 23, 2025. 
Over this interval, temperature increases during the daytime, cools overnight, and rises again the following day, while dewpoint remains comparatively stable, indicating that much of the variation is associated with heating and cooling rather than large shifts in moisture content. 
Relative humidity varies inversely with temperature, wind speed increases during the afternoon with several gusts above 20 kt, visibility remains near 10 statute miles for most of the period, and altimeter pressure changes gradually rather than erratically. 
This figure shows that the ASOS dataset captures coherent meteorological relationships across variables at fine temporal resolution and supports the use of these predictors in feature engineering for the XGBoost model.

```{r, warning=FALSE, echo=TRUE}
#| fig-width: 14
#| fig-height: 20
#| dpi: 300
#| label: fig-multivariable-window
#| fig-cap: "Multi-variable 5-minute meteorological observations at KLAX from July 22 to July 23, 2025. The figure shows short-term co-variation in temperature, dewpoint, relative humidity, wind speed and gusts, visibility, and altimeter pressure over a 36-hour period."


# Select 36 hour time frame
start_time <- ymd_hms("2025-07-22 00:00:00", tz = "UTC")
end_time   <- ymd_hms("2025-07-23 12:00:00", tz = "UTC")

event_data <- last_5_years %>%
  filter(timestamp >= start_time, timestamp <= end_time) %>%
  mutate(
    temp_c = as.numeric(temp_c),
    dewpoint_c = as.numeric(dewpoint_c),
    rh_pct = as.numeric(rh_pct),
    wind_speed_kt = as.numeric(wind_speed_kt),
    wind_gust_kt = as.numeric(wind_gust_kt),
    visibility_sm = as.numeric(visibility_sm),
    altimeter_inhg = as.numeric(altimeter_inhg)
  )

plot_title <- paste0(
  "KLAX 5-minute observations  ",
  format(min(event_data$timestamp), "%H:%M %m/%d/%Y"),
  "  to  ",
  format(max(event_data$timestamp), "%H:%M %m/%d/%Y"),
  " (UTC)"
)

common_x_scale <- scale_x_datetime(
  limits = c(start_time, end_time),
  date_breaks = "6 hours",
  date_labels = "%b %d %H:%M"
)

common_theme <- theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", size = 11),
    axis.title.x = element_blank()
  )

# Panel 1: Temperature and dewpoint
p1 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = temp_c, color = "Temperature"), linewidth = 0.5, na.rm = TRUE) +
  geom_line(aes(y = dewpoint_c, color = "Dewpoint"), linewidth = 0.5, na.rm = TRUE) +
  labs(title = "Temperature (deg C)", y = NULL) +
  scale_color_manual(
    values = c("Temperature" = "firebrick", "Dewpoint" = "darkolivegreen4")
  ) +
  common_x_scale +
  common_theme

# Panel 2: Relative humidity
p2 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = rh_pct, color = "RH"), linewidth = 0.5, na.rm = TRUE) +
  labs(title = "Relative humidity (%)", y = NULL) +
  scale_color_manual(
    values = c("RH" = "steelblue4")
  ) +
  common_x_scale +
  common_theme

# Panel 3: Wind speed and gusts
p3 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = wind_speed_kt, color = "Wind speed"), linewidth = 0.4, na.rm = TRUE) +
  geom_point(
    data = event_data %>% filter(!is.na(wind_gust_kt) & wind_gust_kt > 0),
    aes(y = wind_gust_kt, color = "Gust"),
    size = 1.6,
    alpha = 0.9,
    na.rm = TRUE
  ) +
  labs(title = "Wind (kt)", y = NULL) +
  scale_color_manual(
    values = c("Wind speed" = "grey35", "Gust" = "darkorange")
  ) +
  common_x_scale +
  common_theme

# Panel 4: Visibility
p4 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = visibility_sm, color = "Visibility"), linewidth = 0.5, na.rm = TRUE) +
  labs(title = "Visibility (sm)", y = NULL) +
  scale_color_manual(
    values = c("Visibility" = "mediumpurple4")
  ) +
  common_x_scale +
  common_theme

# Panel 5: Altimeter
p5 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = altimeter_inhg, color = "Altimeter"), linewidth = 0.4, na.rm = TRUE) +
  labs(title = "Altimeter (inHg)", x = "Time (UTC)", y = NULL) +
  scale_color_manual(
    values = c("Altimeter" = "saddlebrown")
  ) +
  common_x_scale +
  common_theme +
  theme(axis.title.x = element_text())

# Combine panels with shared legend at bottom
multi_plot <- (p1 / p2 / p3 / p4 / p5) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = plot_title,
    theme = theme(
      plot.title = element_text(size = 16, hjust = 0.5)
    )
  ) &
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

multi_plot

ggsave(
  filename = file.path(output_dir, "fig_multivariable_recent36h.png"),
  plot = multi_plot,
  width = 11,
  height = 14,
  dpi = 300
)
```

#### Implications for Modeling

Overall, the exploratory results indicate that the KLAX dataset is well suited for prediction of daily high temperature. 
The data exhibits a strong seasonal and diurnal structure, realistic daily variability, and relatively stable short-term temperature behavior, all of which provide meaningful information for a supervised learning model. 
At the same time, the descriptive statistics in @tbl-descriptive_stats suggest that standard preprocessing steps such as scaling and normalization should be sufficient for the use of the tree-based gradient boosting XGBoost model. 

These observations directly inform later stages of the project. 
Seasonal and diurnal patterns visible in @fig-daily-high-ts, @fig-monthly-boxplot, and @fig-diurnal-cycle motivate the inclusion of time-based predictors, while the stable relationships among temperature, dewpoint, humidity, wind, visibility, and pressure shown in @fig-multivariable-window suggest that the aggregated features may improve predictive performance. 
Consequently, the data exploration stage not only describes the structure of the KLAX record but also provides a practical foundation for preprocessing, feature engineering, and model evaluation in the XGBoost framework.

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
