---
title: "Predicting Daily High Temperature Using XGBoost - Summer 2026 "
subtitle: "Los Angeles International Airport (KLAX)"
author: 
  - name: "Alex Lewis"
  - name: "David Moe"
  - name: "Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction

Accurate prediction of daily high temperatures is a critical component of modern weather forecasting, influencing decision-making across sectors such as energy management, transportation, and public safety. 
Reliable temperature forecasts support efficient resource allocation, including electricity demand planning and climate-sensitive operations in aviation and infrastructure. 
In recent years, advances in data-driven modeling have complemented traditional numerical weather prediction methods, offering improved accuracy and adaptability for localized forecasting tasks. 
Machine learning techniques, particularly ensemble methods, have demonstrated strong performance in capturing nonlinear relationships within meteorological data. 
Among these, Extreme Gradient Boosting (XGBoost) has emerged as a powerful and efficient algorithm for predictive modeling in structured datasets [@Chen_2016].

Recent research has increasingly explored the use of machine learning techniques to enhance weather prediction accuracy, particularly in short-term forecasting applications. 
Ensemble learning methods such as XGBoost have been shown to outperform traditional statistical models and complement numerical weather prediction systems by effectively capturing nonlinear relationships and correcting systematic biases [@DONG2023105579]. 
Prior studies have applied XGBoost to tasks such as precipitation forecasting and bias correction of meteorological outputs, demonstrating improved predictive performance across diverse geographic regions [@DONG2023105579]. 
Additionally, comparative analyses of machine learning algorithms highlight the robustness and computational efficiency of gradient boosting approaches when applied to structured environmental data [@Zhou_2023], [@Singh2023MachineLF]. 
Despite these advances, many studies focus on regional or large-scale forecasting problems, leaving opportunities for more localized, site-specific modeling of daily temperature using high-frequency observational data. 
XGBoost has also been documented in tutorial and use-case contexts for its implementation and interpretability in applied machine learning problems [@xgboost_tutorial].

This study focuses on predicting daily high temperatures at Los Angeles International Airport (LAX) using high-frequency observational data obtained from the National Oceanic and Atmospheric Administration (NOAA) Automated Surface Observing System (ASOS) [@asos_dataset_noaa]. 
The dataset consists of five-minute interval weather measurements, providing a detailed temporal view of atmospheric conditions that can be aggregated to support daily forecasting tasks. 
By leveraging this granular data, the study aims to capture short-term variability and improve the accuracy of daily maximum temperature predictions. 
An XGBoost regression model is developed to model the relationship between meteorological features and daily high temperature, with an emphasis on evaluating its predictive performance in a localized airport environment. 
This approach provides a practical framework for applying machine learning techniques to site-specific temperature forecasting using publicly available observational data.

The primary objective of this study is to develop and evaluate an XGBoost-based model for predicting daily high temperatures at LAX using aggregated ASOS observational data. 
Model performance is assessed using a comprehensive set of regression evaluation metrics, including MAE, MSE, RMSE, $R^2$, MAPE, and explained variance, to provide a thorough assessment of prediction accuracy, error magnitude, and model fit. 
In addition, the study examines feature importance to better understand the contribution of key meteorological variables to model predictions. 
By focusing on a single, well-instrumented location and leveraging high-resolution data, this research aims to demonstrate the applicability of machine learning methods for localized temperature forecasting. 
The findings contribute to the growing body of work on data-driven weather prediction and highlight the potential of XGBoost as a practical tool for operational forecasting tasks, consistent with prior statistical temperature prediction work in California and broader North America [@dias2018statistical].

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization - Example

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...


### Data Exploration and Visualization

***Section Under Construction***

***Work In Progress***

#### Data Sources and Collection Process

The temperature data for this study were obtained from the National Oceanic and Atmospheric Administration (NOAA) National Center for Environmental Information (NCEI) Automated Surface Observing System (ASOS) [@asos_dataset_noaa]. The ASOS network provides automated, high-frequency weather observations from thousands of stations across the United States, including Los Angeles International Airport (KLAX). 

The dataset used in this analysis consists of 5-minute interval observations for KLAX from February 1998 to May 2026. Variables include the date and time in UTC: `datetime_utc`, wind direction: `wind_dir`, wind speed: `wind_speed_kt`, wind gusts: `wind_gust_kt`, wind variation: `wind_var`, visibility: `visibility_sm`, weather conditions: `weather`, sky: `sky`, temperature in celsius: `temp_c`, dewpoint in celsius: `dewpoint_c`, rh something: `rh_pct`, altimeter readings: `altimeter_inhg`. For this capstone, the analysis focuses on the most recent five years of observations to ensure data quality and relevance to current climate conditions while maintaining a manageable dataset size for exploration and modeling.

#### Initial Findings and Insights

***The following is all placeholder text to remind me to fill this in.***

*Initial exploration of the KLAX 5-minute temperature data reveals strong seasonal patterns consistent with the Mediterranean climate of Los Angeles, with warmer summers and cooler winters. The distribution of 5-minute temperature changes is tightly centered around zero, indicating that temperature typically evolves smoothly at this timescale, with occasional larger jumps during frontal transitions or rapid diurnal changes.*

*Summary statistics and visualizations indicate a small but non-negligible proportion of missing values in some variables, particularly in older portions of the dataset. Extreme temperature values appear consistent with historical records for the region, with no obvious instrument failures or clear anomalies after basic filtering.*


#### Unexpected Patterns and Anomalies

*More placeholder text below*

*During exploration, a few notable patterns emerged:*

- *Short periods of flat or constant temperature values, likely corresponding to sensor maintenance or data gaps.*
- *Occasional spikes in temperature change magnitude that may correspond to rapid weather transitions or measurement artifacts.*
- *Slight asymmetry in the distribution of positive versus negative 5-minute temperature changes, suggesting a mild tendency for faster warming than cooling at the 5-minute scale.*

*These observations guide subsequent preprocessing steps, including missing value handling, outlier detection, and feature engineering for the XGBoost model.*

```{r, warning=FALSE, echo=T, message=FALSE}
# Setup
library(arrow)
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)
```

```{r, warning=FALSE, echo=TRUE}
# Data Preparation for KLAX 5-minute ASOS Data

# Inspect column names in the parquet file
parquet_path <- "KLAX_5min_complete.parquet"
df <- read_parquet(parquet_path)

# Print all column names
colnames(df)
```

```{r}
# Manually set column names
ts_col <- "datetime_utc"
temp_col <- "temp_c"

df <- df %>%
  mutate(timestamp = as_datetime(get(ts_col), tz = "UTC"))

# Filter to last 5 years
last_5_years <- df %>%
  filter(timestamp >= (now(tzone = "UTC") - years(5)))

cat("Number of observations after filtering to last 5 years:", nrow(last_5_years), "\n")

last_5_years <- last_5_years %>%
  arrange(timestamp) %>%
  mutate(
    temp = get(temp_col),
    temp_change = temp - lag(temp)
  )

cat("Number of NA temp_change values:", sum(is.na(last_5_years$temp_change)), "\n")
cat("Range of temp_change:", range(last_5_years$temp_change, na.rm = TRUE), "\n")

# Remove first row with NA temp_change
last_5_years <- last_5_years %>%
  filter(!is.na(temp_change))

cat("Number of observations after removing NA temp_change:", nrow(last_5_years), "\n")

# Daily high temperature
daily_high <- last_5_years %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(date) %>%
  summarise(
    daily_high_temp = max(temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(daily_high_temp))

cat("Number of days with daily high temperature:", nrow(daily_high), "\n")

# Diurnal data (last 2 years)
diurnal_data <- last_5_years %>%
  filter(timestamp >= (now(tzone = "UTC") - years(2))) %>%
  mutate(hour = hour(timestamp))

diurnal_avg <- diurnal_data %>%
  group_by(hour) %>%
  summarise(
    avg_temp = mean(temp, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r, warning=FALSE, echo=TRUE}
# Visualization 1: Histogram of 5-minute temperature changes
ggplot(last_5_years, aes(x = temp_change)) +
  geom_histogram(
    binwidth = 0.1,
    fill = "steelblue",
    color = "white",
    alpha = 0.8
  ) +
  scale_x_continuous(breaks = seq(-2, 2, by = 0.5)) +
  labs(
    title = "Histogram of 5-Minute Temperature Changes at KLAX",
    subtitle = "Most 5-minute changes are exactly 0°C; extreme changes are rare outliers",
    x = "5-Minute Temperature Change (°C)",
    y = "Frequency"
  ) +
  theme_minimal() +
  xlim(-3, 3)
```

```{r, warning=FALSE, echo=TRUE}
# Proportion of zero changes and histogram of non-zero changes
zero_changes <- last_5_years %>%
  filter(temp_change == 0)

nonzero_changes <- last_5_years %>%
  filter(temp_change != 0)

prop_zero <- nrow(zero_changes) / nrow(last_5_years)

cat("Proportion of 5-minute intervals with exactly 0°C change:", prop_zero, "\n")
cat("Percentage:", 100 * prop_zero, "%\n")

# Histogram of non-zero changes only
ggplot(nonzero_changes, aes(x = temp_change)) +
  geom_histogram(
    binwidth = 0.2,
    fill = "darkorange",
    color = "white",
    alpha = 0.8
  ) +
  labs(
    title = "Histogram of Non-Zero 5-Minute Temperature Changes at KLAX",
    subtitle = sprintf("Proportion of zero changes: %.1f%%", 100 * prop_zero),
    x = "5-Minute Temperature Change (°C)",
    y = "Frequency"
  ) +
  theme_minimal() +
  xlim(-5, 5)
```

```{r, warning=FALSE, echo=TRUE}
# Visualization 2: Daily high temperature time series (5 years)
ggplot(daily_high, aes(x = date, y = daily_high_temp)) +
  geom_line(color = "darkorange", alpha = 0.8) +
  labs(
    title = "Daily High Temperature at KLAX (Last 5 Years)",
    x = "Date",
    y = "Daily High Temperature (°C)"
  ) +
  theme_minimal() +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y")
```

```{r, warning=FALSE, echo=TRUE}
# Visualization 3: Monthly boxplot of daily high temperature
daily_high_months <- daily_high %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE))

ggplot(daily_high_months, aes(x = month, y = daily_high_temp, fill = month)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  labs(
    title = "Monthly Distribution of Daily High Temperature at KLAX",
    x = "Month",
    y = "Daily High Temperature (°C)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

```{r, warning=FALSE, echo=TRUE}
# Visualization 4: Histogram of daily high temperature
ggplot(daily_high, aes(x = daily_high_temp)) +
  geom_histogram(binwidth = 0.5, fill = "forestgreen", color = "white", alpha = 0.8) +
  labs(
    title = "Histogram of Daily High Temperature at KLAX (Last 5 Years)",
    x = "Daily High Temperature (°C)",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r, warning=FALSE, echo=TRUE}
# Visualization 5: Average diurnal temperature curve (by hour)
ggplot(diurnal_avg, aes(x = hour, y = avg_temp)) +
  geom_line(color = "purple", size = 1.2) +
  geom_point(color = "purple", size = 2) +
  labs(
    title = "Average Diurnal Temperature Curve at KLAX (Last 2 Years)",
    x = "Hour of Day",
    y = "Average Temperature (°C)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 0:23)
```

```{r, warning=FALSE, echo=TRUE}
# Summary statistics and abnormal change flags for 5-minute temperature changes

# Absolute temperature changes
last_5_years <- last_5_years %>%
  mutate(abs_temp_change = abs(temp_change))

# Quantiles of absolute changes
abs_q <- quantile(last_5_years$abs_temp_change, probs = c(0.5, 0.75, 0.9, 0.95, 0.99, 1.0))

# Define abnormal threshold (99th percentile)
abnormal_threshold <- abs_q["99%"]

# Flag abnormal changes
last_5_years <- last_5_years %>%
  mutate(abnormal_change = abs_temp_change > abnormal_threshold)

n_abnormal <- sum(last_5_years$abnormal_change)
n_total <- nrow(last_5_years)
pct_abnormal <- 100 * n_abnormal / n_total

# Summary statistics table
summary_stats <- tibble(
  Statistic = c(
    "Total observations",
    "Mean absolute 5-min change (°C)",
    "Median absolute 5-min change (°C)",
    "75th percentile (°C)",
    "90th percentile (°C)",
    "95th percentile (°C)",
    "99th percentile (°C) [abnormal threshold]",
    "Maximum absolute 5-min change (°C)",
    "Number of abnormal changes",
    "Percentage of abnormal changes (%)"
  ),
  Value = c(
    n_total,
    mean(last_5_years$abs_temp_change, na.rm = TRUE),
    median(last_5_years$abs_temp_change, na.rm = TRUE),
    abs_q["75%"],
    abs_q["90%"],
    abs_q["95%"],
    abs_q["99%"],
    max(last_5_years$abs_temp_change, na.rm = TRUE),
    n_abnormal,
    pct_abnormal
  )
)

summary_stats
```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
