---
title: "Predicting Daily High Temperature Using XGBoost - Summer 2026 "
subtitle: "Los Angeles International Airport (KLAX)"
author: 
  - name: "Alex Lewis"
  - name: "David Moe"
  - name: "Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction

The accurate prediction of daily high temperatures is an important component of modern weather forecasting.
It influences decision-making across various market sectors such as energy management, transportation, and public safety. 
By obtaining reliable temperature forecasts, resource can be efficiently allocated in areas such as electricity demand planning and climate-sensitive operations in aviation and infrastructure. 
In recent years, advances in data-driven modeling have improved upon traditional numerical weather prediction methods, offering improved accuracy and adaptability for localized forecasting tasks [@WANG2023101499]. 
Machine learning techniques, in particular ensemble methods, have demonstrated strong performance in capturing nonlinear and interaction relationships within meteorological data [@Airlangga_2025]. 
Among these, Extreme Gradient Boosting (XGBoost) has emerged as a powerful and efficient algorithm for predictive modeling in structured datasets [@Chen_2016].

Recent research has been exploring the use of various machine learning techniques to enhance weather prediction accuracy, particularly in short-term forecasting applications. 
Ensemble learning methods such as XGBoost have been shown to outperform traditional statistical models and complement numerical weather prediction systems by effectively capturing nonlinear relationships and correcting systematic biases [@DONG2023105579]. 
Prior studies have applied XGBoost to tasks such as precipitation forecasting and bias correction of meteorological outputs, demonstrating improved predictive performance across diverse geographic regions [@DONG2023105579]. 
Additionally, a comparative analyses of machine learning algorithms highlights the robustness and computational efficiency of gradient boosting approaches when applied to structured environmental data [@Zhou_2023], [@Singh2023MachineLF], [@Airlangga_2025]. 
Despite these advances, many studies focus on regional or large-scale forecasting problems, leaving opportunities for more localized, site-specific modeling of daily temperature using high-frequency observational data. 
XGBoost has also been documented in tutorial and use-case contexts for its implementation and interpretability in applied machine learning problems [@xgboost_tutorial].

This study focuses on predicting daily high temperatures at the Los Angeles International Airport (KLAX) station using high-frequency observational data obtained from the National Oceanic and Atmospheric Administration (NOAA) Automated Surface Observing System (ASOS) [@asos_dataset_noaa]. 
The dataset consists of five-minute interval weather measurements, providing a detailed view of atmospheric conditions that can be aggregated to support daily forecasting tasks [@Khan_2025_MultiOutputTemp]. 
By leveraging this granular data, this study aims to capture shorter-term variability and improve the accuracy of daily maximum temperature predictions. 
An XGBoost regression model is developed to model the relationship between meteorological features and daily high temperatures, with an emphasis on evaluating its predictive performance in a localized airport environment.
This approach provides a practical framework for applying machine learning techniques to site-specific temperature forecasting using publicly available data.

The primary objective of this study is to develop and evaluate an XGBoost-based model for predicting daily high temperatures at the KLAX airport in Los Angeles, California using the aggregated NOAA ASOS data. 
The model performance is assessed using a comprehensive set of regression evaluation metrics, including MAE, MSE, RMSE, $R^2$, MAPE, and explained variance, to provide a thorough assessment of prediction accuracy, error magnitude, and model fit. 
In addition, the study examines feature importance to better understand the contribution of key meteorological variables to model predictions. 
By focusing on a single location, KLAX, and leveraging high-frequency data, this research aims to demonstrate the applicability of XGBoost for localized temperature forecasting. 
The findings contribute to the growing body of work on data-driven weather prediction and highlight the potential of XGBoost as a practical tool for operational forecasting tasks, consistent with prior statistical temperature prediction work in California and broader North America [@dias2018statistical].

## Methods

The prediction task is to forecast the daily maximum air temperature, $T_{\max}$, at weather station KLAX from surface observations. 
The relationship between the predictors and $T_{\max}$, is highly interaction‑dependent. 
For example, the cooling associated with west‑to‑east wind gusts varies by hour of the day and season, so the marginal effect of any one variable is not constant across the others. 
As a result, the conditional structure cannot be captured by additive linear terms alone. 
This analysis therefore employs gradient‑boosted regression trees as implemented in XGBoost [@Chen_2016], which are well suited to modeling nonlinear, interacting predictors in tabular meteorological data.

We construct an additive ensemble of $K$ regression trees $f_k$ drawn from the space $\mathcal{F}$,

$$
\hat{y}_i = \sum_{k=1}^{K} f_k(\mathbf{x}_i), \qquad f_k \in \mathcal{F}, \tag{1}
$$

where $x_i$ denotes the predictor vector for observation $i$ and $\hat{y}_i$ is the corresponding model prediction.
The ensemble is fitted by stagewise minimization of a regularized objective that penalizes both the number of leaves and the magnitude of leaf prediction scores in each tree:

$$
\mathcal{L} = \sum_{i} l(\hat{y}_i, y_i) + \sum_{k=1}^{K} \left[ \gamma T_k + \tfrac{1}{2}\lambda \lVert w_k \rVert^2 \right]. \tag{2}
$$
where $l(\hat{y}_i, y_i)$ is the loss function (squared error in this study), $T_k$ is the number of leaves in tree $k$, and $w_k$ is the vector of leaf scores for that tree.

This method suits our data for three reasons. 
First, it does not require the predictors to share common units or scales; variables such as wind speed and air temperature can interact naturally within the trees. 
Second, XGBoost includes explicit handling of missing values, so gaps in the observational record, i.e., sensor failures or unmeasured gusts, can be accommodated without discarding entire days or imputing values. 
Third, the fitted model can report which conditions contributed most to its predictions through feature importance and related diagnostics, allowing us to verify that the inferred drivers of high temperatures align with established meteorological cause‑and‑effect relationships.

## Analysis and Results

### Data Exploration and Visualization

#### Data Sources and Collection Process

The temperature data used in this study were obtained from the National Oceanic and Atmospheric Administration (NOAA) National Centers for Environmental Information (NCEI) Automated Surface Observing System (ASOS) network [@asos_dataset_noaa]. 
ASOS is the primary automated surface observing network in the United States and provides continuous meteorological observations at airport-based stations, including Los Angeles International Airport (KLAX), with archived one-minute and five-minute products available through NCEI. 
The dataset used in this study consists of 5-minute interval observations for KLAX from January 1, 2000 through June 7, 2026 and includes variables as described in the following table:

| Variable Name      | Variable       | Description                                                                             |
| ------------------ | -------------- | --------------------------------------------------------------------------------------- |
| Timestamp          | `datetime_utc` | Date and time of the observation in Coordinated Universal Time (UTC).                   |
| Wind Direction     | `wind_dir`     | Direction the wind is coming from, usually reported in degrees.                         |
| Wind Speed         | `wind_speed_kt`| Sustained wind speed in knots.                                                          |
| Wind Gust          | `wind_gust_kt` | Maximum gust wind speed in knots.                                                       |
| Visibility         | `visibility_sm`| Horizontal visibility in statute miles. 10sm Max.                                       |
| Weather            | `weather`      | Reported weather phenomena, such as rain, fog, or thunderstorms.                        |
| Sky Condition      | `sky_condition`| Description of cloud cover or sky state, such as clear, scattered, broken, or overcast. |
| Temperature        | `temp_c`       | Air temperature in degrees Celsius.                                                     |
| Dew Point          | `dewpoint_c`   | Dew point temperature in degrees Celsius.                                               |
| Relative Humidity  | `rh_pct`       | Relative humidity expressed as a percentage.                                            |
| Altimeter Pressure | `altimeter_inhg`| Altimeter setting or atmospheric pressure in inches of mercury.                        |

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Setup Packages
# =========================================================
# Load the packages used for data import, transformation,
# visualization, table output, and panel composition.

library(arrow)
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)
library(patchwork)
library(scales)
library(tibble)

# =========================================================
# Setup Paths
# =========================================================
# Define the source file path, figure output directory, and
# fixed reference date used to construct the recent 5-year window.

parquet_path <- "KLAX_5min_complete.parquet"
output_dir <- "output"
reference_time <- ymd_hms("2026-06-07 23:25:00", tz = "UTC")

if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# =========================================================
# Import Data
# =========================================================
# Read the full KLAX parquet file and standardize the variable
# types needed for tables, figures, and derived calculations.

ts_col <- "datetime_utc"
temp_col <- "temp_c"

df_raw <- read_parquet(parquet_path)

df_all <- df_raw %>%
  mutate(
    timestamp      = as_datetime(.data[[ts_col]], tz = "UTC"),
    temp           = as.numeric(.data[[temp_col]]),
    dewpoint_c     = as.numeric(dewpoint_c),
    rh_pct         = as.numeric(rh_pct),
    wind_dir_deg   = as.numeric(wind_dir),
    wind_speed_kt  = as.numeric(wind_speed_kt),
    wind_gust_kt   = as.numeric(wind_gust_kt),
    visibility_sm  = as.numeric(visibility_sm),
    altimeter_inhg = as.numeric(altimeter_inhg)
  ) %>%
  mutate(
    visibility_sm = ifelse(is.finite(visibility_sm), visibility_sm, NA_real_)
  ) %>%
  mutate(
    wind_gust_kt = ifelse(!is.na(wind_gust_kt) & wind_gust_kt > 100, NA_real_, wind_gust_kt),
    visibility_sm = ifelse(!is.na(visibility_sm) & visibility_sm > 10, NA_real_, visibility_sm)
  ) %>%
  arrange(timestamp)

# =========================================================
# Derive Variables
# =========================================================
# Create the short-term and 3-hour temperature-change fields,
# plus wind-direction groupings used in the distribution analyses.

df_all <- df_all %>%
  mutate(
    temp_change_5min = temp - lag(temp),
    abs_temp_change_5min = abs(temp_change_5min),
    timestamp_lag_3h = lag(timestamp, 36),
    temp_lag_3h = lag(temp, 36),
    temp_change_3h = temp - temp_lag_3h,
    timestamp_future_3h = lead(timestamp, 36),
    temp_future_3h = lead(temp, 36),
    temp_change_next_3h = temp_future_3h - temp,
    wind_group = case_when(
      wind_dir_deg >= 210 & wind_dir_deg <= 330 ~ "West winds",
      wind_dir_deg >= 30 & wind_dir_deg <= 150 ~ "East winds",
      TRUE ~ NA_character_
    ),
    gust_wind_group = case_when(
      wind_dir_deg >= 210 & wind_dir_deg <= 330 ~ "Westerly gusts (west to east)",
      wind_dir_deg >= 30 & wind_dir_deg <= 150 ~ "Easterly gusts (east to west)",
      TRUE ~ NA_character_
    )
  )

# =========================================================
# Build Subsets (Full Record)
# =========================================================
# Create full-record subsets for climatology and temperature-change
# analyses. Exploratory work uses the full 2000–2026 KLAX record.

df_full <- df_all %>%
  filter(!is.na(temp))

df_5min_change_all <- df_all %>%
  filter(!is.na(temp_change_5min))

df_3h_change_all <- df_all %>%
  filter(
    !is.na(temp_change_3h),
    as.numeric(timestamp - timestamp_lag_3h, units = "mins") == 180
  )

df_3h_gust_change <- df_all %>%
  filter(
    !is.na(wind_gust_kt),
    wind_gust_kt < 100,
    !is.na(temp_change_next_3h),
    !is.na(gust_wind_group),
    as.numeric(timestamp_future_3h - timestamp, units = "mins") == 180,
    abs(temp_change_next_3h) <= 10
  )

zero_changes <- df_5min_change_all %>%
  filter(temp_change_5min == 0)

nonzero_changes <- df_5min_change_all %>%
  filter(temp_change_5min != 0)

prop_zero <- nrow(zero_changes) / nrow(df_5min_change_all)

# =========================================================
# Build Daily High Data
# =========================================================
# Build daily high, monthly, and day-of-year summary datasets
# for the visualizations.

daily_high <- df_full %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(date) %>%
  summarise(
    daily_high_temp = max(temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(daily_high_temp))

daily_high_months <- daily_high %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE))

daily_high_doy <- daily_high %>%
  mutate(
    year = year(date),
    doy = yday(date)
  ) %>%
  filter(doy <= 365)

doy_stats <- daily_high_doy %>%
  group_by(doy) %>%
  summarise(
    mean_high = mean(daily_high_temp, na.rm = TRUE),
    q1_high = quantile(daily_high_temp, 0.25, na.rm = TRUE),
    q3_high = quantile(daily_high_temp, 0.75, na.rm = TRUE),
    min_high = min(daily_high_temp, na.rm = TRUE),
    max_high = max(daily_high_temp, na.rm = TRUE),
    .groups = "drop"
  )

day_of_year_month_starts <- tibble(
  doy = c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
  month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
)

# =========================================================
# Build Diurnal Data
# =========================================================
# Compute the average hourly temperature cycle using the same
# 5-year window used for the recent-climatology figures.

diurnal_avg <- df_full %>%
  mutate(hour = hour(timestamp)) %>%
  group_by(hour) %>%
  summarise(
    avg_temp = mean(temp, na.rm = TRUE),
    .groups = "drop"
  )

# =========================================================
# Build Event Window Data
# =========================================================
# Extract the selected 36-hour event window for the multi-panel
# figure that highlights short-term co-variation among variables.

start_time <- ymd_hms("2025-07-22 00:00:00", tz = "UTC")
end_time <- ymd_hms("2025-07-23 12:00:00", tz = "UTC")

event_data <- df_all %>%
  filter(timestamp >= start_time, timestamp <= end_time)

# =========================================================
# Helper Functions
# =========================================================
# Define reusable helpers for descriptive-statistics rows and
# consistent figure export across the capstone project.

make_descriptive_row <- function(data, variable, label) {
  x <- data[[variable]]
  
  tibble(
    Variable = label,
    N = sum(!is.na(x)),
    Mean = mean(x, na.rm = TRUE),
    SD = sd(x, na.rm = TRUE),
    Q1 = quantile(x, 0.25, na.rm = TRUE),
    Median = median(x, na.rm = TRUE),
    Q3 = quantile(x, 0.75, na.rm = TRUE),
    Min = min(x, na.rm = TRUE),
    Max = max(x, na.rm = TRUE)
  )
}

save_plot <- function(plot_object, file_name, width = 7, height = 5, dpi = 300) {
  ggsave(
    filename = file.path(output_dir, file_name),
    plot = plot_object,
    width = width,
    height = height,
    dpi = dpi
  )
}

# =========================================================
# Helper Plot Theme
# =========================================================
# Define a common plotting theme so all figures share a
# consistent visual style and manuscript-ready appearance.

theme_capstone <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    plot.caption = element_text(color = "grey50", size = 8),
    legend.position = "bottom"
  )
```

```{r, warning=FALSE, message=FALSE}
#| label: tbl-observations
# =========================================================
# Observation Tables
# =========================================================

processing_stats <- tibble(
  Statistic = c(
    "Total 5-minute observations",
    "Total non-missing temperature values",
    "Days with daily high temperature values"
  ),
  Value = c(
    nrow(df_all),
    sum(!is.na(df_all$temp)),
    nrow(daily_high)
  )
)

knitr::kable(
  processing_stats,
  caption = "Observation counts for KLAX, 2000-2026",
  digits = 2
)
```

For this analysis, we begin by finding the number of observations within the dataset.  
@tbl-observations shows the total number of observations and the total number of non-missing values.
There are 9,337 values for daily high temperatures. 
This provides a large amount of data for this analysis and the modeling of daily high temperatures later in this study.

#### Descriptive Statistics

```{r, warning=FALSE, message=FALSE}
#| label: tbl-descriptive-stats
# =========================================================
# Descriptive Statistics Table
# =========================================================

continuous_summary <- bind_rows(
  make_descriptive_row(df_full, "wind_speed_kt", "Wind speed (kt)"),
  make_descriptive_row(df_full, "wind_gust_kt", "Wind gust (kt)"),
  make_descriptive_row(df_full, "visibility_sm", "Visibility (statute mi)"),
  make_descriptive_row(df_full, "temp", "Air temperature (°C)"),
  make_descriptive_row(df_full, "dewpoint_c", "Dewpoint temperature (°C)"),
  make_descriptive_row(df_full, "rh_pct", "Relative humidity (%)"),
  make_descriptive_row(df_full, "altimeter_inhg", "Altimeter (inHg)")
)

knitr::kable(
  continuous_summary,
  caption = "Descriptive statistics for KLAX 5-minute variables, 2000–2026.",
  digits = 2
)
```

The descriptive statistics of the selected continuous variables in @tbl-descriptive-stats indicate that most predictors are well-behaved but do not strictly follow a normal Gaussian distribution, with moderate skewness, truncation, and differing scales across variables.
This supports the use of a flexible, nonparametric model such as XGBoost. 
Wind speed and gusts show positive, right-skew with the mean being greater than the median and larger max values further indicating the long right tail.
Wind gusts also show significantly less observations compared to the other variables.
On the other hand, relative humidity exhibits a negative, left-skew with occasional very dry conditions. 

Furthermore, visibility is heavily concentrated at its upper bound of 10 statute miles.
Altimeter readings are tightly clustered around the mean with minimal variance. 
In combination, these features form a predictor set with mixed distributional shapes, bounded ranges, and differing levels of measurement frequency, such as the smaller number of observations for gusts, making tree-based gradient boosting particularly suitable because it does not rely on linearity or normality assumptions and handles non-linear interactions and threshold effects naturally.

Variables such as wind speed, dewpoint temperature, and relative humidity carry substantial variability which can provide strong predictive signals. 
XGBoost can also accommodate the reduced sample size for wind gusts.
Taken together, these descriptive statistics justify the choice of XGBoost as the selected model for this dataset, as it can exploit non-linear relationships and interaction effects while remaining robust to skewed, bounded, and partially observed variables.

```{r, warning=FALSE, message=FALSE}
#| label: tbl-temp-by-wind-group
# =========================================================
# Temperatures by Wind Groups Table
# =========================================================
# Summarize air temperature by east-wind and west-wind groups
# across the full record to match the wind-regime histogram section.

temp_by_wind_summary <- df_all %>%
  filter(!is.na(wind_group), !is.na(temp)) %>%
  group_by(wind_group) %>%
  summarise(
    N = sum(!is.na(temp)),
    Mean = mean(temp, na.rm = TRUE),
    SD = sd(temp, na.rm = TRUE),
    Q1 = quantile(temp, 0.25, na.rm = TRUE),
    Median = median(temp, na.rm = TRUE),
    Q3 = quantile(temp, 0.75, na.rm = TRUE),
    Min = min(temp, na.rm = TRUE),
    Max = max(temp, na.rm = TRUE),
    .groups = "drop"
  )

names(temp_by_wind_summary)[1] <- "Wind Group"

knitr::kable(
  temp_by_wind_summary,
  caption = "Descriptive statistics for air temperature grouped by east-wind and west-wind observations at KLAX, 2000-2026.",
  digits = 2
)
```

@tbl-temp-by-wind-group provides a summary of air temperatures based on whether the wind is blowing from east to west or west to east.
First, west winds more occur frequently with roughly three times as many observations as east winds.
West winds also exhibit higher mean and median temperatures when compared to east winds.
East winds show a much lower minimum temperature, indicating that the east winds are more likely to occur during cooler seasonal periods.
While this appears counterintuitive to the typical understanding that cool marine west winds provide cooler temperatures, this could be indicative of underlying seasonal weather patterns where east winds might prevail during cooler seasons like the winter months.
Furthermore, while the average temperatures may be higher while west winds prevail, wind gusts from the west can still provide a stronger and more frequent cooling effect when compared to east gusts.
This is seen later in this study in @fig-3h-wind-change and supports the idea of an underlying seasonal component for the difference between average temperatures between east and west wind groups.

```{r, warning=FALSE, message=FALSE}
#| label: tbl-temp-summary
# =========================================================
# Temperature Summary Statistics
# =========================================================
# Summarize the magnitude of full-record 5-minute temperature
# changes and identify the 99th-percentile abnormal threshold.

abs_q <- quantile(
  df_5min_change_all$abs_temp_change_5min,
  probs = c(0.5, 0.75, 0.9, 0.95, 0.99, 1.0),
  na.rm = TRUE
)

abnormal_threshold <- abs_q["99%"]

df_5min_change_all <- df_5min_change_all %>%
  mutate(abnormal_change = abs_temp_change_5min > abnormal_threshold)

summary_stats <- tibble(
  Statistic = c(
    "Total observations",
    "Mean absolute 5-minute change (°C)",
    "Median absolute 5-minute change (°C)",
    "75th percentile (°C)",
    "90th percentile (°C)",
    "95th percentile (°C)",
    "99th percentile (°C) [abnormal threshold]",
    "Maximum absolute 5-minute change (°C)",
    "Number of abnormal changes",
    "Percentage of abnormal changes (%)"
  ),
  Value_raw = c(
    nrow(df_5min_change_all),
    mean(df_5min_change_all$abs_temp_change_5min, na.rm = TRUE),
    median(df_5min_change_all$abs_temp_change_5min, na.rm = TRUE),
    abs_q["75%"],
    abs_q["90%"],
    abs_q["95%"],
    abs_q["99%"],
    max(df_5min_change_all$abs_temp_change_5min, na.rm = TRUE),
    sum(df_5min_change_all$abnormal_change, na.rm = TRUE),
    100 * mean(df_5min_change_all$abnormal_change, na.rm = TRUE)
  )
) %>%
  mutate(
    Value = case_when(
      Statistic %in% c("Total observations", "Number of abnormal changes") ~ comma(Value_raw, accuracy = 1),
      Statistic == "Percentage of abnormal changes (%)" ~ number(Value_raw, accuracy = 0.01),
      TRUE ~ number(Value_raw, accuracy = 0.01)
    )
  )


knitr::kable(
  summary_stats %>% select(Statistic, Value),
  caption = "Summary statistics for 5-minute temperature changes at KLAX."
)
```

***ADD DISCUSSION***

#### Initial Findings and Insights

***ADD DISCUSSION***
Exploratory analysis of the filtered KLAX dataset reveals a clear and persistent annual cycle in daily high temperature. 
The time series of daily high temperatures over the last five years in @fig-daily-high-ts shows regular seasonal oscillation, with warmer peaks in late summer and cooler conditions in winter, while still exhibiting day-to-day fluctuations and occasional extreme warm events.

```{r, warning=FALSE, message=FALSE}
#| label: fig-daily-high-ts
#| fig-cap: "Daily high temperature at KLAX over the recent 5-year analysis window."
# =========================================================
# Daily High Time Series Plot
# =========================================================
# Plot the recent 5-year daily high temperature time series.


p_daily_high <- ggplot(daily_high, aes(x = date, y = daily_high_temp)) +
  geom_line(color = "darkorange", alpha = 0.8, linewidth = 0.4) +
  labs(
    title = "Daily High Temperature at KLAX (Full Record)",
    x = "Date",
    y = "Daily High Temperature (°C)"
  ) +
  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y",
    expand = c(0.01, 0.01)
  ) +
  theme_capstone +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


p_daily_high
save_plot(p_daily_high, "fig_daily_high_temp_ts.png")
```

***ADD DISCUSSION***
This seasonal structure is reinforced by the monthly boxplots in @fig-monthly-boxplot, which show median daily highs increasing gradually from winter through late summer and declining again through autumn, with August and September containing the highest medians and some of the warmest outlying days in the sample.

```{r, warning=FALSE, message=FALSE}
#| label: fig-monthly-boxplot
#| fig-cap: "Monthly distribution of daily high temperature at KLAX over the recent 5-year analysis window."
# =========================================================
# Monthly Distribution Boxplot
# =========================================================
# Plot the monthly distribution of daily high temperatures.


p_boxplot <- ggplot(daily_high_months, aes(x = month, y = daily_high_temp, fill = month)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  labs(
    title = "Monthly Distribution of Daily High Temperature at KLAX",
    x = "Month",
    y = "Daily High Temperature (°C)"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    plot.caption = element_text(color = "grey50", size = 8)
  )

p_boxplot
save_plot(p_boxplot, "fig_boxplot.png")
```

***ADD DISCUSSION***
@fig-daily-high-ranges below shows the same seasonal structure at finer resolution, with mean daily highs climbing steadily from January into late summer, peaking in early September, and then declining through autumn into December. 
The shaded band and dashed lines highlight a wider spread between the daily min and max during the cooler winter season, indicating that while the average highs are higher in late summer, the cooler individual winter days actually exhibit greater variability and more frequent extremes than in the warmer months.

```{r, warning=FALSE, message=FALSE}
#| label: fig-daily-high-ranges
#| fig-cap: "Daily high temperature at KLAX by day of year over the recent 5-year analysis window, showing mean, minimum, maximum, and interquartile range values."
# =========================================================
# Daily High Ranges
# =========================================================
# Plot the day-of-year annual cycle using mean, minimum,
# maximum, and interquartile-range summaries.


p_daily_high_ranges <- ggplot(doy_stats, aes(x = doy)) +
  geom_ribbon(
    aes(ymin = q1_high, ymax = q3_high),
    fill = "skyblue3",
    alpha = 0.35
  ) +
  geom_line(
    aes(y = min_high, color = "Min"),
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = max_high, color = "Max"),
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = mean_high, color = "Mean"),
    linewidth = 0.9
  ) +
  scale_color_manual(
    name = "Statistic",
    values = c("Min" = "gray50", "Max" = "gray50", "Mean" = "firebrick3")
  ) +
  scale_x_continuous(
    breaks = day_of_year_month_starts$doy,
    labels = day_of_year_month_starts$month,
    expand = c(0.01, 0.01)
  ) +
  labs(
    title = "Daily High Temperature at KLAX by Day of Year",
    x = "Month",
    y = "Daily High Temperature (°C)"
  ) +
  theme_capstone +
  theme(panel.grid.minor = element_blank())

p_daily_high_ranges
save_plot(p_daily_high_ranges, "fig_daily_high_ranges.png")
```

***ADD DISCUSSION***
The overall distribution of daily high temperatures in @fig-daily-high-hist appears to be normally distributed and concentrated within a moderate range, with most days falling between approximately 16 °C and 24 °C and relatively few very cool or very hot days. 

```{r, warning=FALSE, message=FALSE}
#| label: fig-daily-high-hist
#| fig-cap: "Histogram of daily high temperatures at KLAX over the recent 5-year analysis window."
# =========================================================
# Daily High Histogram
# =========================================================
# Plot the frequency distribution of daily high temperatures.


p_hist_daily <- ggplot(daily_high, aes(x = daily_high_temp)) +
  geom_histogram(binwidth = 1, fill = "forestgreen", color = "white", alpha = 0.8) +
  labs(
    title = "Histogram of Daily High Temperature at KLAX (Recent 5 Years)",
    x = "Daily High Temperature (°C)",
    y = "Frequency"
  ) +
  theme_capstone

p_hist_daily
save_plot(p_hist_daily, "fig_hist_daily_high_temps.png")
```

***ADD DISCUSSION***
Together, @fig-daily-high-ts, @fig-monthly-boxplot, and @fig-daily-high-hist show that the dataset contains strong seasonal signal without being dominated by erratic or implausible daily values, making it suitable for subsequent machine learning analysis.

```{r, warning=FALSE, message=FALSE}
#| label: fig-temp-change-all
#| fig-cap: "Histogram of all 5-minute temperature changes at KLAX across the full observation record."
# =========================================================
# 5 Minute Temp Change All Plot
# =========================================================
# Plot the full-record histogram of all 5-minute temperature changes.


p_hist_change <- ggplot(df_5min_change_all, aes(x = temp_change_5min)) +
  geom_histogram(
    binwidth = 0.1,
    fill = "steelblue",
    color = "white",
    alpha = 0.8
  ) +
  scale_x_continuous(breaks = seq(-2.5, 2.5, by = 0.5)) +
  coord_cartesian(xlim = c(-3, 3)) +
  labs(
    title = "Histogram of 5-Minute Temperature Changes at KLAX",
    subtitle = "Most 5-minute changes are exactly 0°C; larger changes are comparatively rare",
    x = "5-Minute Temperature Change (°C)",
    y = "Frequency"
  ) +
  theme_capstone

p_hist_change
save_plot(p_hist_change, "fig_5min_temp_change_hist.png")
```

***ADD DISCUSSION***
At shorter timescales, the five-minute observations exhibit strong persistence. 
The histogram of all 5-minute temperature changes in @fig-temp-change-all shows an overwhelming concentration at 0 °C, and the non-zero histogram in @fig-temp-change-nonzero confirms that even when change occurs, most changes are small in magnitude and cluster tightly around zero. 

```{r, warning=FALSE, message=FALSE}
#| label: fig-temp-change-nonzero
#| fig-cap: "Histogram of non-zero 5-minute temperature changes at KLAX across the full observation record."
# =========================================================
# 5 Minute Temp Change Excluding Zeros Plot
# =========================================================
# Plot the full-record histogram of non-zero 5-minute changes.


p_non_zero_hist <- ggplot(nonzero_changes, aes(x = temp_change_5min)) +
  geom_histogram(
    binwidth = 0.1,
    fill = "darkorange",
    color = "white",
    alpha = 0.8
  ) +
  scale_x_continuous(breaks = seq(-2.5, 2.5, by = 0.5)) +
  coord_cartesian(xlim = c(-3, 3)) +
  labs(
    title = "Histogram of Non-Zero 5-Minute Temperature Changes at KLAX",
    subtitle = sprintf("Proportion of exact zero changes: %.1f%%", 100 * prop_zero),
    x = "5-Minute Temperature Change (°C)",
    y = "Frequency"
  ) +
  theme_capstone

p_non_zero_hist
save_plot(p_non_zero_hist, "fig_5min_nonzero_change_hist.png")
```

***ADD DISCUSSION***
This pattern is consistent with the summary statistics in @tbl-temp-summary: approximately 70.76% of five-minute intervals show exactly 0 °C change, the mean absolute five-minute change is 0.17 °C, the median absolute change is 0 °C, and the 75th percentile is only 0.5 °C, indicating that short-term temperature changes at KLAX are typically gradual rather than abrupt.

```{r, warning=FALSE, message=FALSE}
#| label: fig-3h-temp-change
#| fig-cap: "Distribution of 3-hour temperature changes from KLAX 5-minute observations across the full record, 2000-2026."
# =========================================================
# 3-hour Temp Changes Histogram
# =========================================================
# Plot the full-record distribution of 3-hour temperature changes.


p_3h_temp_change <- ggplot(df_3h_change_all, aes(x = temp_change_3h)) +
  geom_histogram(binwidth = 1, boundary = 0, fill = "steelblue", colour = "white") +
  scale_x_continuous(breaks = seq(-10, 10, by = 2)) +
  coord_cartesian(xlim = c(-10, 10)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Histogram of 3-Hour Temperature Changes at KLAX",
    subtitle = "Distribution of 3-hour temperature changes from KLAX 5-minute observations, 2000-2026.",
    x = "Temperature Change over 3 Hours (°C)",
    y = "Number of 3-Hour Periods",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station"
  ) +
  theme_capstone

p_3h_temp_change
save_plot(p_3h_temp_change, "fig_3h_temp_change.png")
```

***ADD DISCUSSION***
Given the small changes in temperature during a given 5-minute interval, we expand the interval to a 3-hour range to better observe temperature changes.

```{r, warning=FALSE, message=FALSE}
#| label: fig-3h-wind-change
#| fig-cap: "3-hour temperature change following gust observations at KLAX, grouped by easterly and westerly wind regimes across the full record, 2000-2026."
# =========================================================
# 3-hour Wind Gust Temp Changes Histogram
# =========================================================
# Plot 3-hour temperature change after gust observations,
# grouped by easterly and westerly wind regimes.


p_3h_wind_change <- ggplot(df_3h_gust_change, aes(x = temp_change_next_3h, fill = gust_wind_group)) +
  geom_histogram(
    binwidth = 1,
    boundary = 0,
    position = "identity",
    alpha = 0.6,
    colour = "white"
  ) +
  scale_x_continuous(breaks = seq(-10, 10, by = 2)) +
  coord_cartesian(xlim = c(-10, 10)) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(
    values = c(
      "Westerly gusts (west to east)" = "lightblue",
      "Easterly gusts (east to west)" = "red"
    )
  ) +
  labs(
    title = "Temperature Change After Gusts at KLAX: Onshore vs Offshore",
    subtitle = "3-hour temperature change following gust observations, 2000-2026",
    x = "Temperature Change over 3 Hours (°C)",
    y = "Number of Gust Observations",
    fill = NULL
  ) +
  theme_capstone +
  theme(legend.position = "top")

p_3h_wind_change
save_plot(p_3h_wind_change, "fig_3h_wind_change.png")
```

@fig-3h-wind-change compares the 3-hour temperature changes after gusts from both east and west, or onshore versus offshore gusts.
As previously discussed, average temperatures tend be higher during periods when west winds are present.
However, this does not imply that west winds and gusts do not provide a cooling effect.
As seen in @fig-3h-wind-change, Westerly gusts from offshore provide both a more frequent and stronger cooling effect than easterly gusts from onshore.

```{r, warning=FALSE, message=FALSE}
#| label: fig-diurnal-cycle
#| fig-cap: "Average diurnal temperature cycle at KLAX based on the recent 5-year analysis window."
# =========================================================
# Diurnal Cycle Plot
# =========================================================
# Plot the average diurnal temperature cycle for the recent
# 5-year analysis window.


p_diurnal <- ggplot(diurnal_avg, aes(x = hour, y = avg_temp)) +
  geom_line(color = "purple", linewidth = 1.2) +
  geom_point(color = "purple", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Average Diurnal Temperature Cycle at KLAX (Recent 5 Years)",
    x = "Hour of Day",
    y = "Average Temperature (°C)"
  ) +
  theme_capstone

p_diurnal
save_plot(p_diurnal, "fig_diurnal_cycle.png")
```

***ADD DISCUSSION***
The diurnal cycle plot in @fig-diurnal-cycle provides an additional view of the short-term structure by averaging temperature across each hour of the day during the last two years. That figure shows coolest temperatures during the early morning hours, a steady increase through the morning, and a peak in the early afternoon before gradual evening cooling, which is consistent with expected solar-heating dynamics. This smooth and interpretable daily pattern suggests that time-based predictors such as hour, day, or season are likely to be useful in downstream modeling.

#### Unexpected Patterns and Anomalies

***ADD DISCUSSION***
Although most short-term temperature changes are small, the exploratory analysis also reveals a small number of unusually large jumps. 
Based on the absolute change distribution shown in @fig-temp-change-all and @fig-temp-change-nonzero, the 90th and 95th percentiles are approximately 0.6 °C, while the 99th percentile is 1.1 °C and can be used as a practical threshold for identifying abnormal five-minute changes. Only 4,332 observations, or about 0.99% of the five-minute records, exceed this threshold, indicating that abrupt short-term temperature jumps are rare in the dataset overall.

***ADD DISCUSSION***
The largest observed absolute five-minute change reaches 22.8 °C, which is physically extreme for such a short interval and likely reflects either a reporting artifact, a timing inconsistency, or another unusual observational circumstance rather than typical atmospheric behavior. 
Even so, the rarity of these values suggests that they do not dominate the series, but they do warrant attention during preprocessing because they may affect derived features or distort model fitting if left untreated. 
Furthermore, the presence of long stretches of unchanged temperature and a small number of large jumps indicates that a more practical approach of analysis could be to use a larger timescale.


```{r, warning=FALSE, message=FALSE}
#| fig-width: 14
#| fig-height: 20
#| dpi: 300
#| label: fig-multivariable-window
#| fig-cap: "Multi-variable 5-minute meteorological observations at KLAX from July 22 to July 23, 2025."
# =========================================================
# Multivariable Window
# =========================================================
# Plot a five-panel event-window figure showing the short-term
# co-variation of temperature, moisture, wind, visibility, and pressure.

plot_title <- paste0(
  "KLAX 5-minute observations ",
  format(min(event_data$timestamp), "%H:%M %m/%d/%Y"),
  " to ",
  format(max(event_data$timestamp), "%H:%M %m/%d/%Y"),
  " (UTC)"
)

common_x_scale <- scale_x_datetime(
  limits = c(start_time, end_time),
  date_breaks = "6 hours",
  date_labels = "%b %d %H:%M"
)

common_theme <- theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", size = 11),
    axis.title.x = element_blank()
  )

p1 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = temp, color = "Temperature"), linewidth = 0.5, na.rm = TRUE) +
  geom_line(aes(y = dewpoint_c, color = "Dewpoint"), linewidth = 0.5, na.rm = TRUE) +
  labs(title = "Temperature (°C)", y = NULL) +
  scale_color_manual(values = c("Temperature" = "firebrick", "Dewpoint" = "darkolivegreen4")) +
  common_x_scale +
  common_theme

p2 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = rh_pct, color = "RH"), linewidth = 0.5, na.rm = TRUE) +
  labs(title = "Relative humidity (%)", y = NULL) +
  scale_color_manual(values = c("RH" = "steelblue4")) +
  common_x_scale +
  common_theme

p3 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = wind_speed_kt, color = "Wind speed"), linewidth = 0.4, na.rm = TRUE) +
  geom_point(
    data = event_data %>% filter(!is.na(wind_gust_kt) & wind_gust_kt > 0),
    aes(y = wind_gust_kt, color = "Gust"),
    size = 1.6,
    alpha = 0.9,
    na.rm = TRUE
  ) +
  labs(title = "Wind (kt)", y = NULL) +
  scale_color_manual(values = c("Wind speed" = "grey35", "Gust" = "darkorange")) +
  common_x_scale +
  common_theme

p4 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = visibility_sm, color = "Visibility"), linewidth = 0.5, na.rm = TRUE) +
  labs(title = "Visibility (sm)", y = NULL) +
  scale_color_manual(values = c("Visibility" = "mediumpurple4")) +
  common_x_scale +
  common_theme

p5 <- ggplot(event_data, aes(x = timestamp)) +
  geom_line(aes(y = altimeter_inhg, color = "Altimeter"), linewidth = 0.4, na.rm = TRUE) +
  labs(title = "Altimeter (inHg)", x = "Time (UTC)", y = NULL) +
  scale_color_manual(values = c("Altimeter" = "saddlebrown")) +
  common_x_scale +
  common_theme +
  theme(axis.title.x = element_text())


multi_plot <- (p1 / p2 / p3 / p4 / p5) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = plot_title,
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5))
  ) &
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

multi_plot
save_plot(multi_plot, "fig_multivariable_recent36h.png", width = 11, height = 14)
```

***ADD DISCUSSION***
A more detailed example of short-term multi-variable structure is shown in @fig-multivariable-window, which presents a 36-hour sequence of 5-minute KLAX observations from July 22 to July 23, 2025. 
Over this interval, temperature increases during the daytime, cools overnight, and rises again the following day, while dewpoint remains comparatively stable, indicating that much of the variation is associated with heating and cooling rather than large shifts in moisture content. 
Relative humidity varies inversely with temperature, wind speed increases during the afternoon with several gusts above 20 kt, visibility remains near 10 statute miles for most of the period, and altimeter pressure changes gradually rather than erratically. 
This figure shows that the ASOS dataset captures coherent meteorological relationships across variables at fine temporal resolution and supports the use of these predictors in feature engineering for the XGBoost model.

#### Implications for Modeling

***ADD DISCUSSION***
Overall, the exploratory results indicate that the KLAX dataset is well suited for prediction of daily high temperature. 
The data exhibits a strong seasonal and diurnal structure, realistic daily variability, and relatively stable short-term temperature behavior, all of which provide meaningful information for a supervised learning model. ***Add Discussion on time of day*** 
At the same time, the descriptive statistics in @tbl-descriptive-stats suggest that standard preprocessing steps such as scaling and normalization should be sufficient for the use of the tree-based gradient boosting XGBoost model. 

***ADD DISCUSSION***
These observations directly inform later stages of the project. 
Seasonal and diurnal patterns visible in @fig-daily-high-ts, @fig-monthly-boxplot, and @fig-diurnal-cycle motivate the inclusion of time-based predictors, while the stable relationships among temperature, dewpoint, humidity, wind, visibility, and pressure shown in @fig-multivariable-window suggest that the aggregated features may improve predictive performance. 
Consequently, the data exploration stage not only describes the structure of the KLAX record but also provides a practical foundation for preprocessing, feature engineering, and model evaluation in the XGBoost framework.

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Baseline XGBoost Modeling Dataset
# =========================================================
# Build a daily-level dataset for predicting daily high temperature.

library(xgboost)

# Daily aggregates for key predictors over full record
daily_features <- df_full %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(date) %>%
  summarise(
    daily_high_temp       = max(temp, na.rm = TRUE),
    daily_mean_temp       = mean(temp, na.rm = TRUE),
    daily_mean_dewpoint   = mean(dewpoint_c, na.rm = TRUE),
    daily_mean_rh         = mean(rh_pct, na.rm = TRUE),
    daily_mean_wind_speed = mean(wind_speed_kt, na.rm = TRUE),
    daily_max_wind_gust   = max(wind_gust_kt, na.rm = TRUE),
    daily_mean_visibility = mean(visibility_sm, na.rm = TRUE),
    daily_mean_altimeter  = mean(altimeter_inhg, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(
        daily_high_temp,
        daily_mean_temp,
        daily_mean_dewpoint,
        daily_mean_rh,
        daily_mean_wind_speed,
        daily_max_wind_gust,
        daily_mean_visibility,
        daily_mean_altimeter
      ),
      ~ ifelse(is.infinite(.x), NA_real_, .x)
    )
  ) %>%
  filter(!is.na(daily_high_temp))

daily_features <- daily_features %>%
  mutate(
    daily_mean_temp       = ifelse(daily_mean_temp < -50 | daily_mean_temp > 60, NA_real_, daily_mean_temp),
    daily_mean_dewpoint   = ifelse(daily_mean_dewpoint < -50 | daily_mean_dewpoint > 80, NA_real_, daily_mean_dewpoint),
    daily_mean_rh         = ifelse(daily_mean_rh < 0 | daily_mean_rh > 110, NA_real_, daily_mean_rh),
    daily_mean_wind_speed = ifelse(daily_mean_wind_speed < 0 | daily_mean_wind_speed > 100, NA_real_, daily_mean_wind_speed),
    daily_max_wind_gust   = ifelse(daily_max_wind_gust < 0 | daily_max_wind_gust > 100, NA_real_, daily_max_wind_gust),
    daily_mean_visibility = ifelse(daily_mean_visibility < 0 | daily_mean_visibility > 10, NA_real_, daily_mean_visibility),
    daily_mean_altimeter  = ifelse(daily_mean_altimeter < 25 | daily_mean_altimeter > 32, NA_real_, daily_mean_altimeter)
  )

# Add time-based features
daily_model <- daily_features %>%
  mutate(
    year = year(date),
    doy  = yday(date),
    month = month(date)
  ) %>%
  arrange(date)
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Train / Validation Split (Time-based)
# =========================================================

# Choose a split year, e.g., train up to 2018, validate on 2019+
split_year <- 2018

train_data <- daily_model %>%
  filter(year <= split_year)

valid_data <- daily_model %>%
  filter(year > split_year)

# Feature matrix
feature_cols <- c(
  "doy", "month", "year",
  "daily_mean_temp",
  "daily_mean_dewpoint",
  "daily_mean_rh",
  "daily_mean_wind_speed",
  "daily_max_wind_gust",
  "daily_mean_visibility",
  "daily_mean_altimeter"
)

X_train <- as.matrix(train_data[, feature_cols])
y_train <- train_data$daily_high_temp

X_valid <- as.matrix(valid_data[, feature_cols])
y_valid <- valid_data$daily_high_temp

# Create DMatrix objects for XGBoost
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dvalid <- xgb.DMatrix(data = X_valid, label = y_valid)
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Baseline XGBoost Model (Daily High Temp)
# =========================================================

params <- list(
  objective = "reg:squarederror",  # regression
  eta       = 0.1,                 # learning rate
  max_depth = 6,
  subsample = 0.8,
  colsample_bytree = 0.8,
  nthread   = parallel::detectCores() - 1
)

nrounds <- 300

watchlist <- list(
  train = dtrain,
  valid = dvalid
)

set.seed(42)

bst_baseline <- xgb.train(
  params        = params,
  data          = dtrain,
  nrounds       = nrounds,
  watchlist     = watchlist,
  verbose       = 1,
  print_every_n = 25
)

# Baseline performance on validation set
pred_valid <- predict(bst_baseline, dvalid)

baseline_rmse <- sqrt(mean((pred_valid - y_valid)^2))
baseline_mae  <- mean(abs(pred_valid - y_valid))

baseline_rmse
baseline_mae
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Baseline Feature Importance
# =========================================================

importance <- xgb.importance(
  model = bst_baseline,
  feature_names = feature_cols
)

xgb.plot.importance(importance)
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Regression Metrics on Validation Set
# =========================================================

# Ensure numeric vectors
y_actual <- as.numeric(y_valid)
y_pred   <- as.numeric(pred_valid)

# Residuals
residuals <- y_actual - y_pred

# MAE: Mean Absolute Error
mae <- mean(abs(residuals), na.rm = TRUE)

# MSE: Mean Squared Error
mse <- mean(residuals^2, na.rm = TRUE)

# RMSE: Root Mean Squared Error
rmse <- sqrt(mse)

# R^2: Coefficient of determination
ss_res <- sum(residuals^2, na.rm = TRUE)
ss_tot <- sum((y_actual - mean(y_actual, na.rm = TRUE))^2, na.rm = TRUE)
r_squared <- 1 - ss_res / ss_tot

# MAPE: Mean Absolute Percentage Error
nonzero_idx <- which(y_actual != 0 & !is.na(y_actual) & !is.na(y_pred))
mape <- mean(
  abs((y_actual[nonzero_idx] - y_pred[nonzero_idx]) / y_actual[nonzero_idx]) * 100,
  na.rm = TRUE
)

# Explained variance score
var_actual <- var(y_actual, na.rm = TRUE)
var_res    <- var(residuals, na.rm = TRUE)
explained_variance <- 1 - var_res / var_actual


metrics <- tibble(
  Metric = c("MAE", "MSE", "RMSE", "R_squared", "MAPE_percent", "Explained_variance"),
  Value  = c(mae, mse, rmse, r_squared, mape, explained_variance)
)

print(metrics)
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Validation Diagnostics Data
# =========================================================

valid_diag <- valid_data %>%
  mutate(
    actual_high = as.numeric(y_valid),
    pred_high   = as.numeric(pred_valid),
    residual    = actual_high - pred_high
  )
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Predicted vs Actual Daily Highs
# =========================================================

plot_pred_vs_actual <- ggplot(valid_diag, aes(x = pred_high, y = actual_high)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(
    title    = "Predicted vs Actual Daily High Temperature",
    subtitle = "Validation period (year > split_year)",
    x        = "Predicted daily high (°C)",
    y        = "Actual daily high (°C)"
  ) +
  theme_capstone

plot_pred_vs_actual
save_plot(plot_pred_vs_actual, "xgb_pred_vs_actual_valid.png")
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Residuals over Time
# =========================================================

plot_residual_time <- ggplot(valid_diag, aes(x = date, y = residual)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(alpha = 0.4, color = "darkorange") +
  labs(
    title    = "Residuals over Time (Validation)",
    subtitle = "Actual - predicted daily high (°C)",
    x        = "Date",
    y        = "Residual (°C)"
  ) +
  theme_capstone

plot_residual_time
save_plot(plot_residual_time, "xgb_residuals_time_valid.png")
```

```{r, warning=FALSE, message=FALSE}
# =========================================================
# Residuals vs Actual Temperature
# =========================================================

plot_resid_vs_actual <- ggplot(valid_diag, aes(x = actual_high, y = residual)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(alpha = 0.4, color = "seagreen") +
  labs(
    title    = "Residuals vs Actual Daily High (Validation)",
    subtitle = "Check error structure across temperature range",
    x        = "Actual daily high (°C)",
    y        = "Residual (°C)"
  ) +
  theme_capstone

plot_resid_vs_actual
save_plot(plot_resid_vs_actual, "xgb_residuals_vs_actual_valid.png")
```


### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
