---
title: "Predicting Daily High Temperature Using XGBoost - Summer 2026 "
subtitle: "Los Angeles International Airport (KLAX)"
author: 
  - name: "Alex Lewis"
  - name: "David Moe"
  - name: "Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction {.smaller}

- Daily high temperature forecasting is useful for energy-demand planning, aviation, transportation, and public safety [@WANG2023101499].

- Operational decisions are usually made from site-specific weather conditions rather than broad regional observations.

- Most forecasting studies emphasize broader regional prediction problems, leaving room for local daily temperature prediction to be explored [@Khan_2025_MultiOutputTemp; @dias2018statistical].


## Introduction Continued {.smaller}

- Ensemble models have improved weather forecasting by capturing nonlinear relationships in meteorological data [@Airlangga_2025].

- XGBoost has shown strong performance in weather-related applications such as precipitation forecasting [@DONG2023105579].

- This study asks: **Can XGBoost accurately predict daily high temperatures at KLAX using 5-minute NOAA ASOS observations?**


## Introduction Continued {.smaller}

- Dataset: 5-minute NOAA ASOS observations from KLAX [@asos_dataset_noaa].

- Method: An XGBoost regression model is developed to predict daily high temperature [@Chen_2016].

- Performance is evaluated using MAE, MSE, RMSE, $R^2$, MAPE, and explained variance, and feature importance is analyzed to interpret the model [@Zhou_2023; @Singh2023MachineLF; @xgboost_tutorial].


## Methods

- Target: Daily maximum air temperature, $T_{max}$, at KLAX

- Model: XGBoost regression trees

- XGBoost is well suited for tabular meteorological data:

  - Models nonlinear effects and interactions among predictors.
  
  - Handles missing values from sensor malfunctions without imputation.


## Methods Continued {.smaller}

-   Prediction is the sum of many boosted regression trees.

$$
\hat{y}_i = \sum_{k=1}^{K} f_k(\mathbf{x}_i), \qquad f_k \in \mathcal{F} \tag{1}
$$

- Each tree contributes part of the final temperature prediction.

- The model is trained to balance fit and complexity through regularization.

$$
\mathcal{L} = \sum_{i} l(\hat{y}_i, y_i) + \sum_{k=1}^{K} \left[ \gamma T_k + \tfrac{1}{2}\lambda \lVert w_k \rVert^2 \right] \tag{2}
$$

- The loss term measures prediction error, while the penalty term discourages overly complex trees.

\* *XGBoost Model [@Chen_2016]*


## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
