---
title: "Predicting Daily High Temperature Using XGBoost - Summer 2026 "
subtitle: "Los Angeles International Airport (KLAX)"
author: 
  - name: "Alex Lewis"
  - name: "David Moe"
  - name: "Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction  {.smaller}

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.


## Introduction {.smaller}

-   Predict daily high temperature at KLAX

-   Relevant to aviation and local operations

-   Site-specific forecasting is still difficult

-   NOAA ASOS provides high-frequency weather data at 5-minute intervals

-   XGBoost captures nonlinear weather relationships


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.


## Methods

-   Target: daily high temperature at KLAX

-   Model: XGBoost regression trees

-   Captures nonlinear predictor interactions

-   Handles missing weather observations

-   Feature importance aids interpretation


## Methods Continued

-   Prediction = sum of boosted trees

$$
\hat{y}_i = \sum_{k=1}^{K} f_k(\mathbf{x}_i), \qquad f_k \in \mathcal{F} \tag{1}
$$

-   Objective = fit + regularization

$$
\mathcal{L} = \sum_{i} l(\hat{y}_i, y_i) + \sum_{k=1}^{K} \left[ \gamma T_k + \tfrac{1}{2}\lambda \lVert w_k \rVert^2 \right] \tag{2}
$$

## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
