---
title: "Predicting Daily High Temperature Using XGBoost"
subtitle: "Los Angeles International Airport (KLAX)"
author: 
  - name: "Alex Lewis"
  - name: "David Moe"
  - name: "Advisor: Dr. Cohen"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction {.smaller}

- Daily high temperature forecasting is useful for energy-demand planning, aviation, transportation, and public safety [@WANG2023101499].

- Operational decisions are usually made from site-specific weather conditions rather than broad regional observations.

- Most forecasting studies emphasize broader regional prediction problems, leaving room for local daily temperature prediction to be explored [@Khan_2025_MultiOutputTemp; @dias2018statistical].


## Introduction Continued {.smaller}

- Ensemble models have improved weather forecasting by capturing nonlinear relationships in meteorological data [@Airlangga_2025].

- XGBoost has shown strong performance in weather-related applications such as precipitation forecasting [@DONG2023105579].

- This study asks: **Can XGBoost accurately predict daily high temperatures at KLAX using 5-minute NOAA ASOS observations?**


## Introduction Continued {.smaller}

- **Dataset:** National Oceanic and Atmospheric Administration (NOAA) Automated Surface Observing System (ASOS) [@asos_dataset_noaa]

  - Location: Los Angeles International Airport (KLAX) station
  
  - 5-minute interval weather measurements 

- **Method:** An XGBoost regression model is developed to predict daily high temperature [@Chen_2016].

- **Metrics:** Performance is evaluated using MAE, MSE, RMSE, $R^2$, MAPE, and explained variance, and feature importance is analyzed to interpret the model [@Zhou_2023; @Singh2023MachineLF; @xgboost_tutorial].


## Methods {.smaller}

- **Target:** Daily maximum air temperature, $T_{max}$, at KLAX

- **Model:** XGBoost regression trees

- XGBoost is well suited for tabular meteorological data:

  - Models nonlinear effects and interactions among predictors.
  
  - Handles missing values from sensor malfunctions without imputation.


## Methods Continued {.smaller}

-   Prediction is the sum of many boosted regression trees [@Chen_2016].

$$
\hat{y}_i = \sum_{k=1}^{K} f_k(\mathbf{x}_i), \qquad f_k \in \mathcal{F} \tag{1}
$$

- The model is trained to balance fit and complexity through regularization.

$$
\mathcal{L} = \sum_{i} l(\hat{y}_i, y_i) + \sum_{k=1}^{K} \left[ \gamma T_k + \tfrac{1}{2}\lambda \lVert w_k \rVert^2 \right] \tag{2}
$$

- The loss term measures prediction error, while the penalty term discourages overly complex trees.


## Data Exploration and Visualization {.smaller}
### Data Source

- Obtained from National Oceanic and Atmospheric Administration (NOAA) National Centers for Environmental Information (NCEI) Automated Surface Observing System (ASOS) network [@asos_dataset_noaa]

- ASOS is the primary U.S. airport-based surface weather observing network and provides continuous meteorological observations

- This study uses the Los Angeles International Airport (KLAX) station

- Observations were recorded at 5-minute intervals

- **Study period:** January 1, 2000 to June 7, 2026, covering more than 26 years of records


## Data Exploration and Visualization {.smaller}
### Dataset Structure

- Dataset contains 2,523,612 total 5-minute observations from KLAX

- Variables include Wind direction, wind speed, wind gust, visibility, temperature, dew point, relative humidity, and altimeter pressure

- The modeling target is daily high temperature

- Final target series contains 9,336 daily high-temperature values


```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Load Packages
# =========================================================

library(arrow)
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)
library(patchwork)
library(scales)
library(tibble)
library(tidyr)
library(xgboost)

# =========================================================
# Setup Paths
# =========================================================

parquet_path <- "KLAX_5min_complete.parquet"
reference_time <- ymd_hms("2026-06-07 23:25:00", tz = "UTC")


# =========================================================
# Import Data
# =========================================================

ts_col <- "datetime_utc"
temp_col <- "temp_c"

df_raw <- read_parquet(parquet_path)

df_all <- df_raw %>%
  mutate(
    timestamp      = as_datetime(.data[[ts_col]], tz = "UTC"),
    temp           = as.numeric(.data[[temp_col]]),
    dewpoint_c     = as.numeric(dewpoint_c),
    rh_pct         = as.numeric(rh_pct),
    wind_dir_deg   = as.numeric(wind_dir),
    wind_speed_kt  = as.numeric(wind_speed_kt),
    wind_gust_kt   = as.numeric(wind_gust_kt),
    visibility_sm  = as.numeric(visibility_sm),
    altimeter_inhg = as.numeric(altimeter_inhg)
  ) %>%
  mutate(
    visibility_sm = ifelse(is.finite(visibility_sm), visibility_sm, NA_real_)
  ) %>%
  mutate(
    wind_gust_kt = ifelse(!is.na(wind_gust_kt) & wind_gust_kt > 100, NA_real_, wind_gust_kt),
    visibility_sm = ifelse(!is.na(visibility_sm) & visibility_sm > 10, NA_real_, visibility_sm)
  ) %>%
  arrange(timestamp)

# =========================================================
# Derive Variables
# =========================================================

df_all <- df_all %>%
  mutate(
    temp_change_5min = temp - lag(temp),
    abs_temp_change_5min = abs(temp_change_5min),
    timestamp_lag_3h = lag(timestamp, 36),
    temp_lag_3h = lag(temp, 36),
    temp_change_3h = temp - temp_lag_3h,
    timestamp_future_3h = lead(timestamp, 36),
    temp_future_3h = lead(temp, 36),
    temp_change_next_3h = temp_future_3h - temp,
    wind_group = case_when(
      wind_dir_deg >= 210 & wind_dir_deg <= 330 ~ "West winds",
      wind_dir_deg >= 30 & wind_dir_deg <= 150 ~ "East winds",
      TRUE ~ NA_character_
    ),
    gust_wind_group = case_when(
      wind_dir_deg >= 210 & wind_dir_deg <= 330 ~ "Westerly gusts (west to east)",
      wind_dir_deg >= 30 & wind_dir_deg <= 150 ~ "Easterly gusts (east to west)",
      TRUE ~ NA_character_
    )
  )

# =========================================================
# Build Subsets (Full Record)
# =========================================================

df_full <- df_all %>%
  filter(!is.na(temp))

df_5min_change_all <- df_all %>%
  filter(!is.na(temp_change_5min))

df_3h_change_all <- df_all %>%
  filter(
    !is.na(temp_change_3h),
    as.numeric(timestamp - timestamp_lag_3h, units = "mins") == 180
  )

df_3h_gust_change <- df_all %>%
  filter(
    !is.na(wind_gust_kt),
    wind_gust_kt < 100,
    !is.na(temp_change_next_3h),
    !is.na(gust_wind_group),
    as.numeric(timestamp_future_3h - timestamp, units = "mins") == 180,
    abs(temp_change_next_3h) <= 10
  )

zero_changes <- df_5min_change_all %>%
  filter(temp_change_5min == 0)

nonzero_changes <- df_5min_change_all %>%
  filter(temp_change_5min != 0)

prop_zero <- nrow(zero_changes) / nrow(df_5min_change_all)

# =========================================================
# Build Daily High Data
# =========================================================

daily_high <- df_full %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(date) %>%
  summarise(
    daily_high_temp = max(temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(daily_high_temp))

daily_high_months <- daily_high %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE))

daily_high_doy <- daily_high %>%
  mutate(
    year = year(date),
    doy = yday(date)
  ) %>%
  filter(doy <= 365)

doy_stats <- daily_high_doy %>%
  group_by(doy) %>%
  summarise(
    mean_high = mean(daily_high_temp, na.rm = TRUE),
    q1_high = quantile(daily_high_temp, 0.25, na.rm = TRUE),
    q3_high = quantile(daily_high_temp, 0.75, na.rm = TRUE),
    min_high = min(daily_high_temp, na.rm = TRUE),
    max_high = max(daily_high_temp, na.rm = TRUE),
    .groups = "drop"
  )

day_of_year_month_starts <- tibble(
  doy = c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
  month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
)

# =========================================================
# Build Diurnal Data
# =========================================================

diurnal_avg <- df_full %>%
  mutate(hour = hour(timestamp)) %>%
  group_by(hour) %>%
  summarise(
    avg_temp = mean(temp, na.rm = TRUE),
    .groups = "drop"
  )

# =========================================================
# Build Event Window Data
# =========================================================

start_time <- ymd_hms("2025-07-22 00:00:00", tz = "UTC")
end_time <- ymd_hms("2025-07-23 12:00:00", tz = "UTC")

event_data <- df_all %>%
  filter(timestamp >= start_time, timestamp <= end_time)

# =========================================================
# Helper Functions
# =========================================================

make_descriptive_row <- function(data, variable, label) {
  x <- data[[variable]]
  
  tibble(
    Variable = label,
    N = sum(!is.na(x)),
    Mean = mean(x, na.rm = TRUE),
    SD = sd(x, na.rm = TRUE),
    Q1 = quantile(x, 0.25, na.rm = TRUE),
    Median = median(x, na.rm = TRUE),
    Q3 = quantile(x, 0.75, na.rm = TRUE),
    Min = min(x, na.rm = TRUE),
    Max = max(x, na.rm = TRUE)
  )
}

# =========================================================
# Helper Plot Theme
# =========================================================

theme_capstone <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    plot.caption = element_text(color = "grey50", size = 8),
    legend.position = "bottom"
  )
```


## Visualizations and Insights {.smaller}

::: {.panel-tabset}

### Daily High Time Series

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Daily High Time Series Plot
# =========================================================

p_daily_high <- ggplot(daily_high, aes(x = date, y = daily_high_temp)) +
  geom_line(color = "darkorange", alpha = 0.8, linewidth = 0.4) +
  labs(
    title = "Daily High Temperature at KLAX, 2000-2026",
    x = "Date",
    y = "Daily High Temperature (°C)",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station"
  ) +
  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y",
    expand = c(0.01, 0.01)
  ) +
  theme_capstone +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

p_daily_high
```


### Monthly Boxplot

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Monthly Distribution Boxplot
# =========================================================

p_boxplot <- ggplot(daily_high_months, aes(x = month, y = daily_high_temp, fill = month)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  labs(
    title = "Monthly Distribution of Daily High Temperature at KLAX",
    x = "Month",
    y = "Daily High Temperature (°C)",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40"),
    plot.caption = element_text(color = "grey50", size = 8,),
    legend.position = "none"
  )

p_boxplot
```


### Daily High Ranges

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Daily High Ranges
# =========================================================

p_daily_high_ranges <- ggplot(doy_stats, aes(x = doy)) +
  geom_ribbon(
    aes(ymin = q1_high, ymax = q3_high),
    fill = "skyblue3",
    alpha = 0.35
  ) +
  geom_line(
    aes(y = min_high, color = "Min"),
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = max_high, color = "Max"),
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = mean_high, color = "Mean"),
    linewidth = 0.9
  ) +
  scale_color_manual(
    name = "Statistic",
    values = c("Min" = "gray50", "Max" = "gray50", "Mean" = "firebrick3")
  ) +
  scale_x_continuous(
    breaks = day_of_year_month_starts$doy,
    labels = day_of_year_month_starts$month,
    expand = c(0.01, 0.01)
  ) +
  labs(
    title = "Daily High Temperature at KLAX by Day of Year",
    x = "Month",
    y = "Daily High Temperature (°C)",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station"
  ) +
  theme_capstone +
  theme(panel.grid.minor = element_blank())

p_daily_high_ranges
```


:::


## Visualizations and Insights Continued {.smaller}

::: {.panel-tabset}

### Daily High Histogram

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Daily High Histogram
# =========================================================

p_hist_daily <- ggplot(daily_high, aes(x = daily_high_temp)) +
  geom_histogram(binwidth = 1, fill = "forestgreen", color = "white", alpha = 0.8) +
  labs(
    title = "Histogram of Daily High Temperature at KLAX",
    x = "Daily High Temperature (°C)",
    y = "Frequency",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station"
  ) +
  theme_capstone

p_hist_daily
```


### 3-Hour Temp Change by Wind Group

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# 3-hour Wind Gust Temp Changes Histogram
# =========================================================

p_3h_wind_change <- ggplot(df_3h_gust_change, aes(x = temp_change_next_3h, fill = gust_wind_group)) +
  geom_histogram(
    binwidth = 1,
    boundary = 0,
    position = "identity",
    alpha = 0.6,
    colour = "white"
  ) +
  scale_x_continuous(breaks = seq(-10, 10, by = 2)) +
  coord_cartesian(xlim = c(-10, 10)) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(
    values = c(
      "Westerly gusts (west to east)" = "lightblue",
      "Easterly gusts (east to west)" = "red"
    )
  ) +
  labs(
    title = "Temperature Change After Gusts at KLAX: Onshore vs Offshore",
    x = "Temperature Change over 3 Hours (°C)",
    y = "Number of Gust Observations",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station",
    fill = NULL
  ) +
  theme_capstone +
  theme(legend.position = "top")

p_3h_wind_change
```


### Diurnal Cycle

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Diurnal Cycle Plot
# =========================================================

p_diurnal <- ggplot(diurnal_avg, aes(x = hour, y = avg_temp)) +
  geom_line(color = "purple", linewidth = 1.2) +
  geom_point(color = "purple", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Average Diurnal Temperature Cycle at KLAX",
    x = "Hour of Day",
    y = "Average Temperature (°C)",
    caption = "Source: NOAA ASOS 5-minute data, KLAX station"
  ) +
  theme_capstone

p_diurnal
```


:::


## Modeling and Results {.smaller}
### Preprocessing 

- **Raw input:** KLAX 5-minute ASOS observations from 2000–2026, including temperature, dewpoint, humidity, wind, visibility, and altimeter pressure.

- **Sorting:** Sorted chronologically to preserve temporal order for training and evaluation.

- **Sensor errors:** Checking for sensor errors or anomalies.

  - Visibility values above 10 statute miles were capped because ASOS visibility is reported as 10SM when visibility is 10 miles or greater.
  
  - Wind gusts above 100 kt were removed as implausible extremes relative to the study objective.

- **Missing-values:** Infinite values were relabeled as missing rather than imputed, given XGBoost’s native handling of missing values in tree models.


## Modeling and Results {.smaller}
### Feature Engineering 

- **Target:** Daily high temperature defined as the maximum 5-minute temperature within each calendar day.

- **Lagged features:** Previous-day high temperature, along side daily summary statistics such as mean, minimum, and standard deviation were created to capture persistence.

- **Same-day forecasting window:** Only observations prior to 11:00 A.M. were used so the model relies on information available before the typical high time window.

- **Night vs. morning windows:** Separate 0–6 A.M. and 6–11 A.M. feature windows captured overnight conditions and morning atmospheric change.

- **Seasonality features:** Year, month, day-of-year terms, sinusoidal seasonality, and seasonal indicators helped the model represent annual structure.


## Modeling and Results {.smaller}
### XGBoost

::: {.panel-tabset}

### Model Choice

- **Model:** XGBoost regression - captures nonlinear relationships and interactions common in meteorological data, while also handling missing values natively.

- **Dataset Split:** The dataset was split by year. There are a total of 9,336 daily high target observations.

  - **Train Split:** 2000–2018, consists of 6,800 (72.84%) observations
  
  - **Evaluation Split:** 2019–2026, consists of 2,536 (27.16%) observations 

- **Model Parameters:** Small learning rate and 800 boosting rounds allowed gradual learning.


### Model Parameters

| Hyperparameter                | Value               |
| -------------------           | ----------------    |
| Objective Function            | `Squared Error`     |
| Learning Rate                 | `0.03`              |
| Subsample of Rows             | `0.8`               |
| Subsample of Columns          | `0.8`               |
| Minimum Child Weight          | `5`                 |
| Number of Rounds              | `800`               |
| Early Stopping                | `50`                |

:::

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Create Time of Day Windows
# =========================================================

df_model_base <- df_all %>%
  mutate(
    date  = as.Date(timestamp),
    hour  = hour(timestamp),
    tod_window = case_when(
      hour >= 0  & hour < 6  ~ "night",
      hour >= 6  & hour < 11 ~ "morning",
      TRUE                   ~ NA_character_  # everything else not used
    )
  )

# =========================================================
# Daily Highs
# =========================================================

daily_target <- df_model_base %>%
  group_by(date) %>%
  summarise(
    daily_high_temp = max(temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    daily_high_temp = ifelse(is.infinite(daily_high_temp), NA_real_, daily_high_temp)
  ) %>%
  filter(!is.na(daily_high_temp)) %>%
  arrange(date) %>%
  mutate(
    lag1_daily_high_temp = lag(daily_high_temp, 1)
  )

# =========================================================
# Lagged Features
# =========================================================

safe_max <- function(x) {
  if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE)
}

safe_min <- function(x) {
  if (all(is.na(x))) NA_real_ else min(x, na.rm = TRUE)
}

safe_mean <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

daily_prev_features <- df_model_base %>%
  group_by(date) %>%
  summarise(
    temp_mean_day = safe_mean(temp),
    temp_min_day  = safe_min(temp),
    temp_sd_day   = sd(temp, na.rm = TRUE),
    dew_mean_day  = safe_mean(dewpoint_c),
    rh_mean_day   = safe_mean(rh_pct),
    wind_mean_day = safe_mean(wind_speed_kt),
    gust_max_day  = safe_max(wind_gust_kt),
    vis_mean_day  = safe_mean(visibility_sm),
    alt_mean_day  = safe_mean(altimeter_inhg),
    .groups = "drop"
  ) %>%
  arrange(date) %>%
  mutate(
    lag1_temp_mean_day = lag(temp_mean_day, 1),
    lag1_temp_min_day  = lag(temp_min_day, 1),
    lag1_temp_sd_day   = lag(temp_sd_day, 1),
    lag1_dew_mean_day  = lag(dew_mean_day, 1),
    lag1_rh_mean_day   = lag(rh_mean_day, 1),
    lag1_wind_mean_day = lag(wind_mean_day, 1),
    lag1_gust_max_day  = lag(gust_max_day, 1),
    lag1_vis_mean_day  = lag(vis_mean_day, 1),
    lag1_alt_mean_day  = lag(alt_mean_day, 1),
    lag1_alt_change    = lag(alt_mean_day, 1) - lag(alt_mean_day, 2)
  ) %>%
  select(date, starts_with("lag1_"))

# =========================================================
# Additional Features
# =========================================================

daily_window_features <- df_model_base %>%
  group_by(date, tod_window) %>%
  summarise(
    temp_mean = safe_mean(temp),           
    dew_mean  = safe_mean(dewpoint_c),
    rh_mean   = safe_mean(rh_pct),
    rh_min    = safe_min(rh_pct),
    rh_max    = safe_max(rh_pct),
    wind_mean = safe_mean(wind_speed_kt),
    gust_max  = safe_max(wind_gust_kt),
    vis_mean  = safe_mean(visibility_sm),
    alt_mean  = safe_mean(altimeter_inhg),
    .groups   = "drop"
  ) %>%
  mutate(
    across(
      -c(date, tod_window),
      ~ ifelse(is.infinite(.x), NA_real_, .x)
    )
  ) %>%
  pivot_wider(
    names_from  = tod_window,
    values_from = c(
      temp_mean,
      dew_mean, rh_mean, rh_min, rh_max,
      wind_mean, gust_max, vis_mean, alt_mean
    ),
    names_glue  = "{.value}_{tod_window}"
  ) %>%
  select(!ends_with("_NA")) %>%
  mutate(
    rh_range_night_morning   = rh_max_morning   - rh_min_night,
    temp_range_night_morning = temp_mean_morning - temp_mean_night,
    wind_change_night_morning = wind_mean_morning - wind_mean_night,
    alt_change_night_morning  = alt_mean_morning  - alt_mean_night
  )

# =========================================================
# Prepare Model Data
# =========================================================

model_data <- daily_target %>%
  left_join(daily_prev_features,   by = "date") %>%
  left_join(daily_window_features, by = "date") %>%
  mutate(
    year  = year(date),
    month = month(date),
    doy   = yday(date),
    doy2 = doy^2,
    doy_sin = sin(2 * pi * doy / 365.25),
    doy_cos = cos(2 * pi * doy / 365.25),
    season = case_when(
      month %in% c(12, 1, 2) ~ "winter",
      month %in% c(3, 4, 5)  ~ "spring",
      month %in% c(6, 7, 8)  ~ "summer",
      TRUE                   ~ "fall"
    )
  ) %>%
  arrange(date) %>%
  filter(
    !is.na(lag1_temp_mean_day),
    !is.na(lag1_daily_high_temp)
  ) %>%
  mutate(
    season = factor(season, levels = c("winter", "spring", "summer", "fall"))
  )

# Create season dummies (winter, spring, summer, fall)
season_dummies <- model.matrix(~ season - 1, data = model_data)

model_data <- bind_cols(
  model_data,
  as_tibble(season_dummies)
)

# =========================================================
# Check Features
# =========================================================

feature_cols <- model_data %>%
  select(-date, -daily_high_temp, -month, -season) %>%
  select(where(is.numeric)) %>%                         
  names()

stopifnot(!"daily_high_temp" %in% feature_cols)
```

```{r, results='hide', warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Model Training
# =========================================================

split_year <- 2018

train_data <- model_data %>% filter(year <= split_year)
valid_data <- model_data %>% filter(year > split_year)

X_train <- as.matrix(train_data[, feature_cols])
y_train <- train_data$daily_high_temp

X_valid <- as.matrix(valid_data[, feature_cols])
y_valid <- valid_data$daily_high_temp

dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dvalid <- xgb.DMatrix(data = X_valid, label = y_valid)

set.seed(42)

params <- list(
  objective = "reg:squarederror",
  eta = 0.03,           
  max_depth = 6,        
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5,
  nthread = parallel::detectCores() - 1
)

model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 800,
  evals = list(train = dtrain, valid = dvalid),
  early_stopping_rounds = 50,
  print_every_n = 50,
  verbose = 1
)

pred_valid <- predict(model, dvalid)
```

## Modeling and Results {.smaller}
### Metrics 


```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Regression Metrics on Validation Set
# =========================================================

# Ensure numeric vectors
y_actual <- as.numeric(y_valid)
y_pred   <- as.numeric(pred_valid)

# Residuals
residuals <- y_actual - y_pred

# MAE: Mean Absolute Error
mae <- mean(abs(residuals), na.rm = TRUE)

# MSE: Mean Squared Error
mse <- mean(residuals^2, na.rm = TRUE)

# RMSE: Root Mean Squared Error
rmse <- sqrt(mse)

# R^2: Coefficient of determination
ss_res <- sum(residuals^2, na.rm = TRUE)
ss_tot <- sum((y_actual - mean(y_actual, na.rm = TRUE))^2, na.rm = TRUE)
r_squared <- 1 - ss_res / ss_tot

# MAPE: Mean Absolute Percentage Error
nonzero_idx <- which(y_actual != 0 & !is.na(y_actual) & !is.na(y_pred))
mape <- mean(
  abs((y_actual[nonzero_idx] - y_pred[nonzero_idx]) / y_actual[nonzero_idx]) * 100,
  na.rm = TRUE
)

# Explained variance score
var_actual <- var(y_actual, na.rm = TRUE)
var_res    <- var(residuals, na.rm = TRUE)
explained_variance <- 1 - var_res / var_actual


metrics <- tibble(
  Metric = c("MAE", "MSE", "RMSE", "R Squared", "MAPE %", "Explained Variance"),
  Value  = c(mae, mse, rmse, r_squared, mape, explained_variance)
)

knitr::kable(
  metrics,
  caption = "Metric Results from XGBoost Model",
  digits = 3
)
```


## Modeling and Results {.smaller}
### Featrue Importance 

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Model Feature Importance Plot
# =========================================================
importance <- xgb.importance(
  model = model,
  feature_names = feature_cols
)

importance_tbl <- as_tibble(importance) %>%
  arrange(desc(Gain)) %>%      
  slice(1:10)

feature_importance_plot <- ggplot(
  importance_tbl,
  aes(x = reorder(Feature, Gain), y = Gain)
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_capstone +              
  labs(
    title = "XGBoost Feature Importance (Top 10)",
    x = "Feature",
    y = "Gain"
  )

feature_importance_plot
```


## Modeling and Results {.smaller}
### Diagnostics 

::: {.panel-tabset}

### Predicted vs. Actual Daily Highs

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Validation Diagnostics Data
# =========================================================

valid_diag <- valid_data %>%
  mutate(
    actual_high = as.numeric(y_valid),
    pred_high   = as.numeric(pred_valid),
    residual    = actual_high - pred_high
  )

# =========================================================
# Predicted vs Actual Daily Highs
# =========================================================

plot_pred_vs_actual <- ggplot(valid_diag, aes(x = pred_high, y = actual_high)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(
    title    = "Predicted vs Actual Daily High Temperature",
    subtitle = "Validation period 2019-2026",
    x        = "Predicted daily high (°C)",
    y        = "Actual daily high (°C)"
  ) +
  theme_capstone

plot_pred_vs_actual
```


### Residuals vs. Actual Daily Highs

```{r, warning=FALSE, echo=F, message=FALSE}
# =========================================================
# Residuals vs Actual Temperature
# =========================================================

plot_resid_vs_actual <- ggplot(valid_diag, aes(x = actual_high, y = residual)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(alpha = 0.4, color = "seagreen") +
  labs(
    title    = "Residuals vs Actual Daily High (Validation)",
    subtitle = "Check error structure across temperature range",
    x        = "Actual daily high (°C)",
    y        = "Residual (°C)"
  ) +
  theme_capstone

plot_resid_vs_actual
```


:::


## Conclusion {.smaller}

- XGBoost predicted KLAX daily highs with high accuracy, usually within about 1 °C.

- The model captured most of the day-to-day temperature variability.

- Morning weather, previous-day highs, humidity, and wind were the most useful predictors.

- The model performed best on typical days and struggled more on very hot or very cold days.

- Adding large-scale weather patterns could improve predictions for extreme events.


## References
